### 2019-07-05 paper/paper.tex
add paper

% JuliaCon Proceedings
% Template with instructions: https://www.overleaf.com/read/dqjbrhqxjpwq

\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{url}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\newcommand\EatDot[1]{}

\begin{document}

\input{header}

\maketitle

\begin{abstract}

\small{\texttt{StateSpaceModels.jl}} is an open-source Julia package for modeling, forecasting and simulating time series in a state-space framework. The package represents a straightforward tool that can be useful for a wide range of applications that deal with time series. In addition, it contains features that are not present in related commercial software, such as Monte Carlo simulation and the possibility of setting any user-defined linear model.

\end{abstract}

\section{Motivation} \label{sec:introduction}

State-space modeling is a classical framework in control engineering that represents a system through the definition of input, state, and output variables \cite{zadeh2008linear}. Input variables are external entities that are inserted into the system and can serve as control inputs or noise. State variables represent unobserved components that evolve through time following a given state equation and also depending on the values of the input variables. Finally, output variables result from the realization of the state plus noise factors and represent the observable outcome of the system. In general, state-space models make use of the Kalman filter \cite{kalman1960new, welch1995introduction} to obtain predictive estimates for the state.

Due to its comprehensive form and wide range of potential applications, state-space models found a niche in time-series modeling, forecasting, and simulation, representing a flexible framework for time-series analysis with time-varying parameters \cite{durbin2012time}. The ability to conveniently define the evolution of a time series allows the characterization of stochastic components, such as trend and seasonality, that are non-trivial to model in other frameworks.

There are several packages focused on state-space models for time-series analysis in other languages, such as \texttt{KFAS} in R \cite{kfas} and \texttt{Statsmodels} in Python \cite{seabold2010statsmodels}, as well as commercial software such as \texttt{STAMP} \cite{koopman2000stamp}. Among related Julia packages, we highlight:
\begin{itemize}
    \item ~\texttt{Kalman.jl} \cite{kalmanjl}, which implements Kalman filtering and smoothing.
    \item ~\texttt{StateSpace.jl} \cite{statespacejl}, a control-oriented package with several versions of the Kalman filter.
    \item ~\texttt{ControlSystems.jl} \cite{controlsystemsjl}, a package aimed at control applications with numerous features, including state-space modeling.
\end{itemize}

The contributions brought by these packages notwithstanding, there are currently no state-space model packages aimed at time-series analysis in Julia. Furthermore, some have not been updated for a long time and are not functional in Julia 1.0. Thus, the main objectives of \texttt{StateSpaceModels.jl} \cite{statespacemodels} are:

\begin{enumerate}
    \item to fill this gap by implementing a general and intuitive framework in Julia for modeling, estimating, forecasting and simulating time series with state-space models;
    \item to provide an open-source package that is capable of performing the same functions as related commercial software such as \texttt{STAMP} and more;
    \item to develop a package for time-series analysis with state-space models that is fully implemented in Julia, contrary to related packages in other languages that depend on C or Fortran \mbox{routines} \cite{kfas}.
\end{enumerate}

The remainder of this paper is organized as follows. In Section \ref{sec:statespace}, the Gaussian state-space framework is introduced. In Section \ref{sec:modeling}, the procedures for model specification are presented. Section \ref{sec:estimation} explains the filtering, estimation and smoothing processes. In Section \ref{sec:forecasting}, the forecasting and simulation procedures are presented. Section \ref{sec:applications} contains examples of applications that illustrate the use of the package. Finally, conclusions and future work are provided in Section \ref{sec:conclusion}

\section{Gaussian state-space framework} \label{sec:statespace}

For the sake of consistency and readability, we utilize in the package the same notation as \cite{durbin2012time}. Suppose we have a series of observations $y_{1}, \dots, y_{n}$. In a state-space framework, it is assumed that the observations depend on a set of unobserved components denominated states, or simply the state, and denoted by $\alpha_{1}, \dots, \alpha_{n}$. The state components often have physical interpretations, such as trend and seasonality in a time-series context or position and speed in a control setting.

The main idea behind state-space modeling is to define the evolution of the state and its relation with the observed variables. This is done through the following two equations, denoted the observation equation and the state equation, respectively:
%
\begin{align}
    y_{t} &= Z_{t} \alpha_{t} + \varepsilon_{t}, \hspace{0.08\columnwidth} \varepsilon_{t} \sim N(0, H_{t}), \label{eq:obs_equation} \\
    \alpha_{t+1} &= T_{t} \alpha_{t} + R_{t} \eta_{t}, \quad \eta_{t} \sim N(0, Q_{t}), \label{eq:state_equation}
\end{align}
%
\noindent where $y_{t}$ is a $p \times 1$ vector of observations and $\alpha_{t}$ is an unobserved $m \times 1$ vector representing the state at instant $t$. Note that the behavior of the system over time is determined by $\alpha_{t}$ as defined in \mbox{Eq. \eqref{eq:state_equation}}, but $\alpha_{t}$ cannot be directly observed, contrary to $y_{t}$.

Matrices $Z_{t}$, $T_{t}$, and $R_{t}$ are the ones that define how the observations relate to the state and how the state evolves over time, and are generally assumed to be known. The error terms $\varepsilon_{t}$ and $\eta_{t}$ are supposed to be serially independent and independent of each other. Finally, $H_{t}$ and $Q_{t}$ are the covariance matrices of the error terms. In this package, it is assumed that $T$, $R$, $H$, and $Q$ are not time-varying, as this is the case for the vast majority of practical applications. Conversely, $Z_{t}$ is allowed to vary over time.

Next, for illustration purposes, we present an example model and show how it can be inserted into the state-space framework. Consider the following model, called the linear trend model:
%
\begin{align}
    y_{t} &= \mu_{t} + \varepsilon_{t}, \hspace{0.115\columnwidth} \varepsilon_{t} \sim N(0, \sigma^{2}_{\varepsilon}), \label{eq:lineartrend1} \\
    \mu_{t+1} &= \mu_{t} + \nu_{t} + \xi_{t}, \quad \xi_{t} \sim N(0, \sigma^{2}_{\xi}), \label{eq:lineartrend2} \\
    \nu_{t+1} &= \nu_{t} + \zeta_{t}, \hspace{0.12\columnwidth} \zeta_{t} \sim N(0, \sigma^{2}_{\zeta}), \label{eq:lineartrend3}
\end{align}
%
\noindent where $\mu_{t}$ represents the trend component and $\nu_{t}$ the slope component. Note that the we can rewrite this model in the framework of Eq. \eqref{eq:obs_equation}--\eqref{eq:state_equation} in the following manner:
%
\begin{align}
    y_{t} &= \begin{bmatrix} 1 & 0 \end{bmatrix} \alpha_{t} + \varepsilon_{t}, \hspace{0.15\columnwidth} \varepsilon_{t} \sim N(0, \sigma^{2}_{\varepsilon}), \\
    \alpha_{t+1} &= \begin{bmatrix} 1 & 1 \\ 0 & 1 \end{bmatrix} \alpha_{t} + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}\!\eta_{t}, \quad \eta_{t} \sim N \bigg(\!\begin{bmatrix} 0 \\ 0 \end{bmatrix}\!, \begin{bmatrix} \sigma^{2}_{\xi} & 0 \\ 0 & \sigma^{2}_{\zeta} \end{bmatrix}\!\bigg).
\end{align}
%
Similarly, more complex and sophisticated models can also be modeled in the state-space framework.

\section{Model specification} \label{sec:modeling}

Model specification is the step wherein the user defines the state-space model they want to consider. This is done through the creation of a \texttt{StateSpaceModel} structure that contains the observations $y_{t}$ as well as matrices $Z_{t}$, $T$, and $R$.

\subsection{Predefined models} \label{subsec:predefined}

A set of predefined classical models is available, namely the local level model, the linear trend model, the basic structural model, and the structural model with exogenous variables. These models can be conveniently defined by calling a function and providing the observations $y_{t}$. Note that all models automatically extend to the multivariate case.

\subsubsection{Local level model}

The local level model consists of a stochastic level component that is defined by a random walk:
%
\begin{align}
    y_{t} &= \mu_{t} + \varepsilon_{t}, \quad \varepsilon_{t} \sim N(0, \sigma^{2}_\varepsilon), \label{eq:locallevel1} \\
    \mu_{t+1} &= \mu_{t} + \xi_{t}, \quad \xi_{t} \sim N(0, \sigma^{2}_\xi). \label{eq:locallevel2}
\end{align}
%
A local level \texttt{StateSpaceModel} can be created using:
%
\begin{lstlisting}[language = Julia]
model = local_level(y)
\end{lstlisting}
%
\noindent where \texttt{y} is an observation vector in the univariate case or a matrix in column-wise fashion in the multivariate case, i.e., each column representing a variable and each line representing a time period.
%

\subsubsection{Linear trend model}

The linear trend model consists of a local level model with the addition of a stochastic slope term, as seen in Eq. \eqref{eq:lineartrend1}--\eqref{eq:lineartrend3}. A linear trend \texttt{StateSpaceModel} can be created using:
%
\begin{lstlisting}[language = Julia]
model = linear_trend(y)
\end{lstlisting}

\subsubsection{Structural model}

The basic structural model consists of three stochastic components: trend, slope, and seasonality, defined in the following manner:
%
\begin{align}
    y_{t} &= \mu_{t} + \gamma_{t} + \varepsilon_{t}, \hspace{0.09\columnwidth} \varepsilon_{t} \sim N(0, \sigma^{2}_\varepsilon), \label{eq:structural1} \\
    \mu_{t+1} &= \mu_{t} + \nu_{t} + \xi_{t}, \hspace{0.095\columnwidth} \xi_{t} \sim N(0, \sigma^{2}_\xi), \label{eq:structural2} \\
    \nu_{t+1} &= \nu_{t} + \zeta_{t}, \hspace{0.18\columnwidth} \zeta_{t} \sim N(0, \sigma^{2}_\zeta), \label{eq:structural3} \\
    \gamma_{t+1} &= \sum_{j=1}^{s-1} \gamma_{t+1-j} + \omega_{t}, \quad \omega_{t} \sim N(0, \sigma^{2}_\omega). \label{eq:structural4}
\end{align}
%
In this model, there is a stochastic trend that contains a slope component, similarly to the linear trend model. Furthermore, a stochastic seasonal component is modeled so that its sum over the seasonality period $s$ is equal to zero except for a noise factor. A basic structural \texttt{StateSpaceModel} can be created using:
%
\begin{lstlisting}[language = Julia]
model = structural(y, s)
\end{lstlisting}

Additionally, a structural model can contain exogenous variables, also known as explanatory variables. Exogenous variables represent external factors that are correlated to the phenomenon in question and can, thus, be used to improve the estimation and forecasting by increasing the available information. This model is identical to \eqref{eq:structural1}--\eqref{eq:structural4} except for the observation equation, which has an additional exogenous factor:
%
\begin{align}
    y_{t} &= \mu_{t} + \gamma_{t} + \theta^{\top} X_{t} + \varepsilon_{t}, \quad \varepsilon_{t} \sim N(0, \sigma^{2}_\varepsilon) \label{eq:structural_exogenous}
\end{align}
%
A structural model with exogenous variables can be created with:
%
\begin{lstlisting}[language = Julia]
model = structural(y, s; X = X)
\end{lstlisting}
%
\noindent where \texttt{X} is a matrix containing the exogenous variables observations in column-wise fashion.

\subsection{User-defined models}

Apart from the well-known predefined models presented in Section \ref{subsec:predefined}, the package allows the input of any user-defined linear model through the definition of matrices $Z_{t}$, $T$, and $R$ through the following constructor:
%
\begin{lstlisting}[language = Julia]
model = StateSpaceModel(y, Z, T, R)
\end{lstlisting}

In the simpler case where $Z_{t}$ is constant, the argument \texttt{Z} can be input as an \texttt{Array\{Float64, 2\}}. Conversely, if $Z_{t}$ is time-varying, it must be input as an \texttt{Array\{Float64, 3\}} where the third dimension represents the time periods.

\section{Filtering, estimation, and smoothing} \label{sec:estimation}

Next, the model needs to be estimated. In this step, values for the covariance matrices $H$ and $Q$, initially assumed to be unknown, are obtained via maximum likelihood estimation \cite{casella2002statistical}. In order to do that, the Kalman filter needs to be utilized. Additionally, a smoother is applied in order to obtain the so-called smoothed estimates for the state.

After specifying the model, the estimation step can be called with the function \texttt{statespace}:
%
\begin{lstlisting}[language = Julia]
ss = statespace(model; verbose = 1)
\end{lstlisting}

In this step, the Kalman filtering, the maximum likelihood estimation of the fixed parameters, and the smoothing are conducted. The optional argument \texttt{verbose}, which defaults to 1, specifies the verbosity: 0 for no output, 1 for some progress information, and 2 for progress information as well as the optimization log.

An example of the estimation log with default verbosity can be seen below.
%
\begin{lstlisting}[language = Julia]
julia> ss = statespace(model)
==============================================================
                  StateSpaceModels.jl v0.2.0
 (c) Raphael Saavedra, Guilherme Bodin, and Mario Souto, 2019
--------------------------------------------------------------
            Starting state-space model estimation.
    Initiating maximum likelihood estimation with 3 seeds.
--------------------------------------------------------------
             Seed 0 is aimed at degenerate cases.
--------------------------------------------------------------
||    seed    |     log-likelihood      |      time (s)     ||
||       0    |        -16217.4939      |          0.21     ||
||       1    |         -1350.4763      |          2.65     ||
||       2    |         -1350.4763      |          4.08     ||
||       3    |         -1350.4763      |          5.69     ||
--------------------------------------------------------------
           Maximum likelihood estimation complete.
                  Log-likelihood: -1350.4763
             End of state-space model estimation.
==============================================================
\end{lstlisting}

\subsection{Filtering}

The filtering step derives the predictive state as well as its covariance matrix at every time period via the Kalman filter, i.e., it estimates
%
\begin{align}
    a_{t+1} &= \mathbb{E}[\alpha_{t+1} | Y_{t}], \\
    P_{t+1} &= \mathbb{V}[\alpha_{t+1} | Y_{t}],
\end{align}
%
where $Y_{t}$ denotes the set of time periods $y_{1}, \dots, y_{t}$.

Furthermore, the filter also computes the innovations, or prediction errors, as well as their covariance matrix at every time period:
%
\begin{align}
    v_{t} &= y_{t} - Z a_{t}, \\
    F_{t} &= \mathbb{V}[v_{t}].
\end{align}
%

The package contains an implementation of the standard Kalman filter and the square-root Kalman filter \cite{durbin2012time}. The latter is a variant which utilizes Cholesky decomposition to ensure that the computed covariance matrices are positive semidefinite, thus avoiding numerical errors.

Additionally, in order to provide a flexible framework, we allow the utilization of any user-implemented variant of the Kalman filter through an abstract type called \texttt{AbstractFilter}:
%
\begin{lstlisting}[language = Julia]
struct MyKalmanFilter <: AbstractFilter
\end{lstlisting}

The choice of the filter type must be done in the estimation step as follows:
%
\begin{lstlisting}[language = Julia]
ss = statespace(model; filter_type = SquareRootFilter)
\end{lstlisting}

If no \texttt{filter\_type} is provided, the standard type \texttt{KalmanFilter} is utilized as a default option. Further information on Kalman filtering can be found in Chapter 4.3 of \cite{durbin2012time}.

\subsection{Estimation}

The Kalman filter is responsible for deriving estimates for the predictive state and for its covariance matrix at each time period, but it is dependent on the values of the noise covariances $H$ and $Q$. Thus, these constant parameters need to be estimated via maximum likelihood \cite{casella2002statistical}. The log-likelihood function is given by
%
\begin{align}
    \ell(Y_{n}) = -\frac{np}{2} \log 2 \pi - \frac{1}{2} \sum_{t=1}^{n} (\log |F_{t}| + v_{t}^{\top} F_{t}^{-1} v_{t}),
\end{align}
%
which depends on the innovations and their covariance matrix at each time period. Therefore, at each optimization iteration, the Kalman filter needs to be executed.

The implemented optimization method \texttt{RandomSeedsLBFGS} generates random initial values for the parameters and then uses the L-BFGS algorithm \cite{liu1989limited} to maximize the log-likelihood function via the unconstrained optimization package \texttt{Optim.jl} \cite{mogensen2018optim}.

Additionally, similar to the filtering step, users are able to define any desired optimization method through abstract type \texttt{AbstractOptimizationMethod}:
%
\begin{lstlisting}[language = Julia]
struct MyOptimizationMethod <: AbstractOptimizationMethod
\end{lstlisting}

As with the filter type, the choice of optimization method must be done in the estimation step:
%
\begin{lstlisting}[language = Julia]
ss = statespace(model; optimization_method = RandomSeedsLBFGS())
\end{lstlisting}

More advanced parameters, such as optimizer tolerances and the number of seeds, can be set within the \texttt{RandomSeedsLBFGS} constructor.

\subsection{Smoothing}

The smoothing step is conducted after estimation is complete. It consists in obtaining the so-called smoothed state and its covariance matrix at each time period, i.e.,
%
\begin{align}
    \hat{\alpha}_{t} &= \mathbb{E}[\alpha_{t} | Y_{n}], \\
    V_{t} &= \mathbb{V}[\alpha_{t} | Y_{n}].
\end{align}
%
Similar to the filtering step, the smoothing is conducted through an iterative process. The procedure is done backwards, starting from the final time period $n$. The smoothed state estimates are useful for analyzing the behavior of the state components. Further information on state smoothing can be found on Chapter 4.4 of \cite{durbin2012time}.

\subsection{Missing observations}

One interesting property of the Kalman filter and smoother is their ability to treat missing observations. Suppose we have a time series spanning time periods $t = 1, \dots, n$, but for some periods $\tau, \dots, \tau^{*}$ we have no observations available. This represents an obstacle for estimating a model in several frameworks. Nonetheless, the Kalman filter and smoother allow the derivation of minimum variance linear unbiased estimates for the missing observations so that a completion of the time series is possible. Conveniently, the only necessary modification in the filtering and smoothing equations is considering $Z_{t} = 0$ for the missing period.

In the package, this is implemented so that any \texttt{NaN} values in the observations are considered missing observations and automatically treated in the filtering and smoothing steps. In this way, the estimation is successfully conducted even when the time series of interest is incomplete.

Consider the following simulated time series composed by a growing trend plus a Gaussian noise. Additionally, we remove observations 10 through 20 to treat them as missing:
%
\begin{lstlisting}[language = Julia]
# Create growing trend series with Gaussian noise
y = collect(1:0.25:20) + 0.5*randn(77)
# Remove observations 10 through 20
y[10:20] .= NaN
# Specify the state-space model and estimate it
model = linear_trend(y)
ss = statespace(model)
\end{lstlisting}
%
The results of the estimation step are presented in Fig. \ref{fig:missing_observations}. The filtered state and smoothed state estimates can be accessed in \texttt{ss.filter.a} and \texttt{ss.smoother.alpha}, respectively. It is clear in Fig. \ref{fig:missing_observations} that the filter and smoother are able to effectively capture the growing linear trend of the series even in the period where observations are missing.
%
\begin{figure}[h]
	\centering
	\includegraphics[width=\columnwidth]{Figures/missing_obs.pdf}
	\caption{Automatic completion of missing observations in the filtering and smoothing.}
	\label{fig:missing_observations}
\end{figure}

\subsection{Estimation outputs}

At the end of the filtering, estimation, and smoothing routine executed by the \texttt{statespace} function, a \texttt{StateSpace} structure is returned. It contains six fields:

\begin{itemize}
    \item ~\texttt{model}: the previously defined \texttt{StateSpaceModel}.
    \item ~\texttt{filter}: a \texttt{FilteredState} structure representing the result of the Kalman filter. It contains the predictive state and the innovations as well as their covariance matrices at each time period. Furthermore, there is a flag indicating if steady state was attained and the time period when it was attained.
    \item ~\texttt{smoother}: a \texttt{SmoothedState} structure representing the result of the smoothing procedure. It contains the smoothed state as well as its covariance matrix at each time period.
    \item ~\texttt{covariance}: a \texttt{StateSpaceCovariance} structure which contains the covariance matrices of the observation and state noises.
    \item ~\texttt{filter\_type}: the defined filter type.
    \item ~\texttt{optimization\_method}: the defined optimization method.
\end{itemize}

\section{Forecasting and Simulation} \label{sec:forecasting}

Finally, after the model has been estimated, forecasting and Monte Carlo simulation can be conducted. In the case of forecasting, estimates for the future values of the time series are computed along with their probability distributions. Alternatively, it is possible to simulate an arbitrary number of future scenarios of the series, which is useful for a wide range of applications such as stochastic optimization.

The minimum mean square error forecasts for $y_{t}$ can be obtained by treating future values of $y_{t}$ as missing values. Forecasting is conducted with the function \texttt{forecast}, which receives a \texttt{StateSpace} structure coming from the estimation step and the number of time periods ahead to be forecast, and outputs the minimum square error forecasts and the predictive distributions of $y_{t}$ at each time period:
%
\begin{lstlisting}[language = Julia]
pred, dist = forecast(ss, N)
\end{lstlisting}
%
Alternatively, another powerful tool is the simulation of future scenarios, since these are used as input by several applications, such as certain stochastic optimization models. Performing Monte Carlo simulation in a time-series state-space framework involves sampling several scenarios for the observation and state errors utilizing the estimated variances $H$ and $Q$ at each time period, and then computing the state-space recursions for each scenario.

Monte Carlo simulation can be conducted with the use of the function \texttt{simulate}, which receives a \texttt{StateSpace} structure coming from the estimation step, the number of time periods ahead $N$, and the number of scenarios $S$ to be simulated, and outputs an $N \times S$ matrix of scenarios for $y_{t}$:
%
\begin{lstlisting}[language = Julia]
scenarios = simulate(ss, N, S)
\end{lstlisting}

\section{Applications} \label{sec:applications}

In this section, we present several practical applications which can be addressed via state-space modeling. We utilize \texttt{StateSpaceModels.jl} to tackle the presented problems and provide the results. Furthermore, for the sake of reproducibility, the code of some of the examples is in the folder \texttt{examples} in the \texttt{StateSpaceModels.jl} repository.

\subsection{Airline passengers}

As a first example, let us use the classical monthly airline passengers time series. In order to avoid multiplicative effects, we use the well-known approach of taking the log of the series. Fig. \ref{fig:log_airline_forecast} shows the log-airline passengers time series. We can estimate a structural \texttt{StateSpaceModel} as follows.
%
\begin{lstlisting}[language = Julia]
# Load the AirPassengers dataset
AP = CSV.read("AirPassengers.csv")
# Take the log of the series
logAP = log.(Vector{Float64}(AP[:Passengers]))
# Specify the state-space model
model = structural(logAP, 12)
# Estimate the state-space model
ss = statespace(model)
\end{lstlisting}

By estimating a structural model, we can analyzed the individual components of the series, such as trend and seasonality. These components are presented in Fig. \ref{fig:log_airline_components} and represent dimensions 1 and 3 of the smoothed state, respectively.

We can also forecast the following two years of the time series by using the \texttt{forecast} function. The result is displayed in Fig. \ref{fig:log_airline_forecast}.
%
\begin{lstlisting}[language = Julia]
# Number of months ahead to be forecast
N = 24
# Perform forecasting
pred, dist = forecast(ss, N)
\end{lstlisting}
%
\begin{figure}[h]
	\centering
	\includegraphics[width=\columnwidth]{Figures/log_airline_forecast.pdf}
	\caption{Log-airline passengers historical data and forecasting.}
	\label{fig:log_airline_forecast}
\end{figure}
%
\begin{figure}[h]
	\centering
	\includegraphics[width=\columnwidth]{Figures/log_airline_components.pdf}
	\caption{Smoothed trend and seasonal components of the log-airline passengers time series.}
	\label{fig:log_airline_components}
\end{figure}

\subsection{Electricity consumption}

Another practical example is the study of monthly electricity consumption in a given area. In this example, we will use real data from a Brazilian distribution company. In addition, we will consider a temperature series, which is highly correlated to electricity consumption in Brazil, as an exogenous variable. A similar study using \texttt{StateSpaceModels.jl} was conducted in \cite{saavedra2018simulating}.

In this case, our objective is to simulate a large set of future scenarios that can be used as inputs in a stochastic optimization problem with the goal of obtaining the best contracting strategy for a distribution company. To this end, the function \texttt{simulate} will be used to perform Monte Carlo simulation. We will also use the square-root Kalman filter for this application.
%
\begin{lstlisting}[language = Julia]
# Specify the state-space model
model = structural(consumption, 12; X = temperature)
# Estimate the state-space model
ss = statespace(model; filter_type = SquareRootFilter)
# Number of months ahead to be simulated
N = 24
# Number of scenarios to be simulated
S = 1000
# Perform simulation
sim = simulate(ss, N, S)
\end{lstlisting}

The time series and the resulting simulation can be seen in Fig. \ref{fig:energy_simulation}. The future scenarios are graphically represented by their mean, which is identical to the forecast at each time period, as well as the 5\% and 95\% quantiles.
%
\begin{figure}[h]
	\centering
	\includegraphics[width=\columnwidth]{Figures/energy_simulation.pdf}
	\caption{Monthly energy consumption historical data and simulation of future scenarios.}
	\label{fig:energy_simulation}
\end{figure}

\subsection{Vehicle tracking}

Finally, in order to illustrate one application that does not fall into any of the predefined models, thus requiring a user-defined model, let us consider an example from control theory. More precisely, we are going to use \texttt{StateSpaceModels.jl} to track a vehicle from noisy sensor data. In this case, $y_t$ is a $2 \times 1$ observation vector representing the corrupted measurements of the vehicle's position on the two-dimensional plane in instant $t$. Since sensors collect the observations with the presence of additive Gaussian noise, we need to filter the observation in order to obtain a better estimate of the vehicle's position.

The position and speed in each dimension compose the state of the vehicle. Let us refer to $x_t^{(d)}$ as the position on the axis $d$ and to $\dot{x}^{(d)}_t$ as the speed on the axis $d$ in instant $t$. Additionally, let $\eta^{(d)}_t$ be the input drive force on the axis $d$, which acts as state noise. For a single dimension, we can describe the vehicle dynamics as 
\begin{equation}
\begin{aligned}
    & x_{t+1}^{(d)} = x_t^{(d)} + \Big( 1 - \frac{\rho \Delta_t}{2} \Big) \Delta_t \dot{x}^{(d)}_t + \frac{\Delta^2_t}{2} \eta_t^{(d)}, \\
    & \dot{x}^{(d)}_{t+1} = (1 - \rho) \dot{x}^{(d)}_{t} + \Delta_t \eta^{(d)}_t,
\end{aligned}\label{eq_control}
\end{equation}
where $\Delta_t$ is the time step and $\rho$ is a known damping effect on speed.

We can cast the dynamical system (\ref{eq_control}) as a state-space model in the following manner:
\begin{align*} 
    y_t &= \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 0 & 1 & 0 \end{bmatrix} \alpha_{t+1} + \varepsilon_t, \\
    \alpha_{t+1} &= \begin{bmatrix} 1 & (1 - \tfrac{\rho \Delta_t}{2}) \Delta_t & 0 & 0 \\ 0 & (1 - \rho) & 0 & 0 \\ 0 & 0 & 1 & (1 - \tfrac{\rho \Delta_t}{2}) \\ 0 & 0 & 0 & (1 - \rho) \end{bmatrix} \alpha_{t} + \begin{bmatrix} \tfrac{\Delta^2_t}{2} & 0 \\ \Delta_t & 0 \\ 0 & \tfrac{\Delta^2_t}{2} \\ 0 & \Delta_t \end{bmatrix} \eta_{t},
\end{align*}
where $\alpha_t = (x_t^{(1)}, \dot{x}^{(1)}_{t}, x_t^{(2)}, \dot{x}^{(2)}_{t})^{\top}$ and $\eta_t = (\eta^{(1)}_t, \eta^{(2)}_t)^{\top}$.

\vspace{0.3cm}

We can formulate the vehicle tracking problem in the \texttt{StateSpaceModels.jl} framework as:

\begin{lstlisting}[language = Julia]
# State transition matrix
T = kron(Matrix{Float64}(I, p, p), [1 (1 - ρ * Δ / 2) * Δ; 0 (1 - ρ * Δ)])
# Input matrix
R = kron(Matrix{Float64}(I, p, p), [.5 * Δ^2; Δ])
# Output (measurement) matrix
Z = kron(Matrix{Float64}(I, p, p), [1 0])
# User defined model
model = StateSpaceModel(y, Z, T, R)
# Estimate vehicle speed and position
ss = statespace(model)
\end{lstlisting}

In this example, we define the noise variances $H$ and $Q$, generate the noises and simulate a random vehicle trajectory using the state-space equations:
%
\begin{lstlisting}[language = Julia]
# Generate random actuators
Q = .5 * Matrix{Float64}(I, q, q)
η = MvNormal(zeros(q), Q)
# Generate random measurement noise
H = 2. * Matrix{Float64}(I, p, p)
ε = MvNormal(zeros(p), H)
# Simulate vehicle trajectory
α = Matrix{Float64}(undef, n + 1, m)
y = Matrix{Float64}(undef, n, p)
for t in 1:n
    y[t, :] = Z * α[t, :] + rand(ε)
    α[t + 1, :] = T * α[t, :] + R * rand(η)  
end
\end{lstlisting}

An illustration of the results can be seen in Fig. \ref{fig:vehicle_tracking}. It can be seen that the measurements are reasonably noisy when compared to the true position. Furthermore, the estimated positions, represented by the smoothed state, effectively estimate the true positions with small inaccuracies. 
%
\begin{figure}[h]
	\centering
	\includegraphics[width=\columnwidth]{Figures/vehicle_tracking.pdf}
	\caption{Vehicle tracking using a state-space model.}
	\label{fig:vehicle_tracking}
\end{figure}

\section{Conclusion} \label{sec:conclusion}

\texttt{StateSpaceModels.jl} is a flexible package for time-series modeling, forecasting, and simulating that is fully implemented in Julia. The package contains an implementation of the Kalman filter and smoother as well as their square-root variants. Additionally, users have the ability to use any implemented filter or optimization method. Missing observations in the form of \texttt{NaN} values are automatically treated in the filtering and smoothing steps.

Besides comprising several predefined classical models, it is also possible to define any linear model with the \texttt{StateSpaceModel} constructor. Forecasting and Monte Carlo simulation of future scenarios are also available. Finally, the package documentation contains a manual with straightforward examples that are simple to reproduce.

\input{bib.tex}

\end{document}

### 2019-07-06 src/kalman.jl
filter with the same results as KFAS

@@ -3,8 +3,16 @@
 
 Kalman filter with big Kappa initialization.
 """
-function kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = 1e-5) where Typ <: AbstractFloat
+function kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}) where Typ <: AbstractFloat
+    # if model.mode == "time-variant"
+    if true
+        return time_variant_kalman(model, H, Q)
+    else
+        return time_invariant_kalman(model, H, Q)
+    end
+end
 
+function time_invariant_kalman(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}) where Typ <: AbstractFloat
     # Load dimensions
     n, p, m, r = size(model)
 
@@ -15,6 +23,8 @@ function kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}; t
     # Predictive state and its covariance
     a = Matrix{Float64}(undef, n+1, m)
     P = Array{Float64, 3}(undef, m, m, n+1)
+    att = Matrix{Float64}(undef, n, m)
+    Ptt = Array{Float64, 3}(undef, m, m, n)
 
     # Innovation and its covariance
     v = Matrix{Float64}(undef, n, p)
@@ -33,38 +43,79 @@ function kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}; t
 
     # Kalman filter
     for t = 1:n
-        # Check for missing values
-        if any(isnan.(y[t, :]))
-            steadystate  = false
-            v[t, :]      = NaN*ones(p)
-            F[:, :, t]   = Z[:, :, t]*P[:, :, t]*Z[:, :, t]' + H
-            K[:, :, t]   = T * P[:, :, t] * Z[:, :, t]' * inv(F[:, :, t])
-            a[t+1, :]    = T*a[t, :]
-            P[:, :, t+1] = ensure_pos_sym(T*P[:, :, t]*T' + R*Q*R')
+        if check_missing_observation(y[t, :])
+            v[t, :]      = fill(NaN, p)
+            F[:, :, t]   = fill(NaN, (p, p))
+            att[t, :]    = a[t, :]
+            Ptt[:, :, t] = P[:, :, t]
[…]

### 2019-07-06 src/structures.jl

@@ -31,8 +31,10 @@ end
 # Auxiliary structure for Kalman filter
 mutable struct KalmanFilter <: AbstractFilter
     a::Matrix{Float64} # predictive state
+    att::Matrix{Float64}
     v::Matrix{Float64} # innovations
     P::Array{Float64, 3} # covariance matrix of the predictive state
+    Ptt::Array{Float64, 3}
     F::Array{Float64, 3} # covariance matrix of the innovations
     steadystate::Bool # flag that indicates if steady state was attained
     tsteady::Int # instant when steady state was attained; in case it wasn't, tsteady = n+1
@@ -187,12 +189,12 @@ Following the notation of on the book \"Time Series Analysis by State Space Meth
 * `tstady` Instant when steady state was attained; in case it wasn't, `tsteady = n+1`
 """
 struct FilteredState
-    a::Matrix{Float64} # predictive state
-    v::Matrix{Float64} # innovations
-    P::Array{Float64, 3} # lower triangular matrix with sqrt-covariance of the predictive state
-    F::Array{Float64, 3} # lower triangular matrix with sqrt-covariance of the innovations
-    steadystate::Bool # flag that indicates if steady state was attained
-    tsteady::Int # instant when steady state was attained; in case it wasn't, tsteady = n+1
+    a::Matrix{Float64} 
+    v::Matrix{Float64} 
+    P::Array{Float64, 3} 
+    F::Array{Float64, 3} 
+    steadystate::Bool 
+    tsteady::Int 
 end
 
 """

### 2019-07-06 src/utils.jl

@@ -50,6 +50,10 @@ function ensure_pos_sym(M::Matrix{T}; ϵ::T = 1e-8) where T <: AbstractFloat
     return (M + M')/2 + ϵ*I
 end
 
+function check_missing_observation(y::Vector{Float64})
+    return any(isnan, y)
+end
+
 function Base.show(io::IO, ss::StateSpace)
     println("An estimated state-space model with")
     println("filter_type = $(ss.filter_type),")

### 2019-07-06 test/runtests.jl

@@ -1,5 +1,6 @@
+push!(LOAD_PATH, "/Users/guilhermebodin/Documents/StateSpaceModels.jl/src")
 using Test, StateSpaceModels, Statistics, CSV, LinearAlgebra
-
+cd("test")
 const SSM = StateSpaceModels
 # Run tests
 include("test_utils.jl")

### 2019-07-06 src/kalman.jl
update

@@ -3,16 +3,17 @@
 
 Kalman filter with big Kappa initialization.
 """
-function kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}) where Typ <: AbstractFloat
-    # if model.mode == "time-variant"
-    if true
+function kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ},
+                       tol::Typ = 1e-5) where Typ <: AbstractFloat
+    if model.mode == "time-variant"
         return time_variant_kalman(model, H, Q)
     else
-        return time_invariant_kalman(model, H, Q)
+        return time_invariant_kalman(model, H, Q, tol)
     end
 end
 
-function time_invariant_kalman(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}) where Typ <: AbstractFloat
+function time_invariant_kalman(model::StateSpaceModel, H::Matrix{Typ}, 
+                               Q::Matrix{Typ}, tol::Typ) where Typ <: AbstractFloat
     # Load dimensions
     n, p, m, r = size(model)
 
@@ -41,28 +42,43 @@ function time_invariant_kalman(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix
     a[1, :]    = zeros(m, 1)
     P[:, :, 1] = 1e6 .* Matrix(I, m, m)
 
+    RQR = R * Q * R'
     # Kalman filter
     for t = 1:n
         if check_missing_observation(y[t, :])
+            steadystate  = false
             v[t, :]      = fill(NaN, p)
             F[:, :, t]   = fill(NaN, (p, p))
             att[t, :]    = a[t, :]
             Ptt[:, :, t] = P[:, :, t]
             a[t+1, :]    = T * att[t, :]
-            P[:, :, t+1] = ensure_pos_sym(T * Ptt[:, :, t] * T' + R*Q*R')
+            P[:, :, t+1] = ensure_pos_sym(T * Ptt[:, :, t] * T' + RQR)
+        elseif steadystate
+            v[t, :]      = y[t, :] - Z[:, :, t] * a[t, :] # v_t = y_t - Z_t * a_t
[…]

### 2019-07-07 test/runtests.jl
Update runtests.jl

@@ -1,7 +1,7 @@
-push!(LOAD_PATH, "/Users/guilhermebodin/Documents/StateSpaceModels.jl/src")
 using Test, StateSpaceModels, Statistics, CSV, LinearAlgebra
-cd("test")
+
 const SSM = StateSpaceModels
+
 # Run tests
 include("test_utils.jl")
 include("test_userdefined.jl")

### 2019-07-07 src/sqrt_kalman.jl
add filtered state

@@ -25,7 +25,7 @@ function sqrt_kalman_filter(model::StateSpaceModel, sqrtH::Matrix{Typ}, sqrtQ::M
 
     # Auxiliary matrices
     U2star = Array{Float64, 3}(undef, m, p, n)
-    
+
     # Steady state initialization
     steadystate  = false
     tsteady      = n+1
@@ -75,6 +75,39 @@ function sqrt_kalman_filter(model::StateSpaceModel, sqrtH::Matrix{Typ}, sqrtQ::M
 
     # Return the auxiliary filter structre
     return SquareRootFilter(a, v, sqrtP, sqrtF, steadystate, tsteady, K)
+end
+
+"""
+    filtered_state(model::StateSpaceModel, sqrt_filter::SquareRootFilter)
+
+Obtain the filtered state estimates and their covariance matrices.
+"""
+function filtered_state(model::StateSpaceModel, sqrt_filter::SquareRootFilter)
+
+    # Load dimensions data
+    n, p, m, r = size(model)
+
+    # Load system data
+    Z, T, R = ztr(model)
+
+    # Load filter data
+    a = sqrt_filter.a
+    v = sqrt_filter.v
+    F = gram_in_time(sqrt_filter.sqrtF)
+    P = gram_in_time(sqrt_filter.sqrtP)
+
+    # Filtered state and its covariance
+    att = Matrix{Float64}(undef, n, m)
+    Ptt = Array{Float64, 3}(undef, m, m, n)
+
+    for t = 1:n
+        PZF = P[:, :, t] * Z[:, :, t]' * inv(F[:, :, t])
+        att[t, :]    = a[t, :] + PZF * v[t, :]
+        Ptt[:, :, t] = ensure_pos_sym(P[:, :, t] - PZF * Z[:, :, t] * P[:, :, t])
+    end
+
+    return att, Ptt
+
 end
 
 """
@@ -151,7 +184,7 @@ function sqrt_smoother(model::StateSpaceModel, sqrt_filter::SquareRootFilter)
     Nstar  = [Z[:, :, 1]' * inv(sqrtF[:, :, 1]) L[:, :, 1]' * sqrtN[:, :, 1]]
     G      = qr(Nstar').Q
     NstarG = Nstar*G
-    
+
     sqrtN_0  = NstarG[1:m, 1:m]
     P_1 = gram(sqrtP[:, :, 1])
     alpha[1, :] = a[1, :] + P_1 * r_0
[…]

### 2019-07-07 src/structures.jl

@@ -1,5 +1,5 @@
-export StateSpaceDimensions, StateSpaceModel, StateSpaceCovariance, SmoothedState, 
-        FilteredState, StateSpace, KalmanFilter, SquareRootFilter, RandomSeedsLBFGS
+export StateSpaceDimensions, StateSpaceModel, StateSpaceCovariance, SmoothedState,
+        FilterOutput, StateSpace, KalmanFilter, SquareRootFilter, RandomSeedsLBFGS
 
 # Abstract types
 """
@@ -25,7 +25,7 @@ mutable struct SquareRootFilter <: AbstractFilter
     sqrtF::Array{Float64, 3} # lower triangular matrix with sqrt-covariance of the innovations
     steadystate::Bool # flag that indicates if steady state was attained
     tsteady::Int # instant when steady state was attained; in case it wasn't, tsteady = n+1
-    K::Array{Float64, 3} # Kalman gain  
+    K::Array{Float64, 3} # Kalman gain
 end
 
 # Auxiliary structure for Kalman filter
@@ -66,7 +66,7 @@ end
 """
     StateSpaceDimensions
 
-StateSpaceModel dimensions, following the notation of on the book 
+StateSpaceModel dimensions, following the notation of on the book
 \"Time Series Analysis by State Space Methods\" (2012) by J. Durbin and S. J. Koopman.
 
 * `n` is the number of observations
@@ -104,7 +104,7 @@ struct StateSpaceModel
     mode::String
 
     function StateSpaceModel(y::Matrix{Float64}, Z::Array{Float64, 3}, T::Matrix{Float64}, R::Matrix{Float64})
-        
+
         # Validate StateSpaceDimensions
         ny, py = size(y)
         pz, mz, nz = size(Z)
@@ -116,7 +116,7 @@ struct StateSpaceModel
         dim = StateSpaceDimensions(ny, py, mr, rr)
         new(y, Z, T, R, dim, "time-variant")
     end
-    
+
     function StateSpaceModel(y::Matrix{Float64}, Z::Matrix{Float64}, T::Matrix{Float64}, R::Matrix{Float64})
 
         # Validate StateSpaceDimensions
@@ -147,15 +147,15 @@ Following the notation of on the book \"Time Series Analysis by State Space Meth
 * `Q` covariance matrix of the state vector ``Q_t``
 """
 struct StateSpaceCovariance
-    H::Matrix{Float64} 
+    H::Matrix{Float64}
     Q::Matrix{Float64}
 
[…]

### 2019-07-07 paper/header.tex
updating paper

% **************GENERATED FILE, DO NOT EDIT**************

\title{StateSpaceModels.jl: a Julia Package for \\ Time-Series Analysis in a State-Space Framework}

\author[1,2]{Raphael Saavedra}
\author[1,2]{Guilherme Bodin}
\author[1,2]{Mario Souto}
\affil[1]{Laboratory of Applied Mathematical Programming and Statistics (LAMPS)}
\affil[2]{Pontifical Catholic University of Rio de Janeiro (PUC-Rio), Brazil}

\keywords{State-space models, time-series analysis, Kalman filter, Monte Carlo simulation, forecasting}

### 2019-07-07 paper/paper.tex
updating paper

@@
 
-The filtering step derives the predictive state as well as its covariance matrix at every time period via the Kalman filter, i.e., it estimates
+The filtering step derives the predictive and filtered states as well as their covariance matrices at every time period via the Kalman filter. The predictive state and its variance are given by
 %
@@
     P_{t+1} &= \mathbb{V}[\alpha_{t+1} | Y_{t}],
+\end{align}
+%
+while the filtered state and its variance are given by
+%
+\begin{align}
+    a_{t|t} &= \mathbb{E}[\alpha_{t} | Y_{t}], \\
+    P_{t|t} &= \mathbb{V}[\alpha_{t} | Y_{t}],
 \end{align}
@@
 \ref{fig:missing_observations}.
-The filtered state and smoothed state estimates can be accessed in \texttt{ss.filter.a} and \texttt{ss.smoother.alpha}, respectively.
+The predictive, filtered, and smoothed state estimates can be accessed in \texttt{ss.filter.a}, \texttt{ss.filter.att}, and \texttt{ss.smoother.alpha}, \mbox{respectively}.
 It is clear in Fig.
@@
     \item ~\texttt{model}: the previously defined \texttt{StateSpaceModel}.
-    \item ~\texttt{filter}: a \texttt{FilteredState} structure representing the result of the Kalman filter.
-It contains the predictive state and the innovations as well as their covariance matrices at each time period.
+    \item ~\texttt{filter}: a \texttt{FilterOutput} structure representing the result of the Kalman filter.
+It contains the predictive and filtered states as well as the innovations and their covariance matrices at each time period.
 Furthermore, there is a flag indicating if steady state was attained and the time period when it was attained.

### 2019-07-07 paper/header.tex
final paper update

@@
 
-\title{StateSpaceModels.jl: a Julia Package for \\ Time-Series Analysis in a State-Space Framework}
+\title{StateSpaceModels.jl: a Julia Package for Time-Series Analysis in a State-Space Framework}
 
-\author[1,2]{Raphael Saavedra}
-\author[1,2]{Guilherme Bodin}
-\author[1,2]{Mario Souto}
+\author[1, 2]{Raphael Saavedra}
+\author[1, 2]{Guilherme Bodin}
+\author[1, 2]{Mario Souto}
 \affil[1]{Laboratory of Applied Mathematical Programming and Statistics (LAMPS)}
@@
 
-\keywords{State-space models, time-series analysis, Kalman filter, Monte Carlo simulation, forecasting}
+\keywords{State-space models, time-series analysis, Kalman filter, forecasting, Monte Carlo simulation}

### 2019-08-05 benchmark/bench.jl
add benchmark and move missing_obsrvations inside the model

@@ -5,29 +5,43 @@ verbose = 0
 
 locallevel = local_level(Matrix{Float64}(locallevel_series))
 bench_local_level = @benchmark statespace($locallevel, filter_type = $SquareRootFilter, verbose = $verbose)
-# 1st August 2019
+# 5th August 2019
 # BenchmarkTools.Trial:
-#   memory estimate:  555.70 MiB
-#   allocs estimate:  2964817
+#   memory estimate:  187.38 MiB
+#   allocs estimate:  2398549
 #   --------------
-#   minimum time:     248.796 ms (23.64% GC)
-#   median time:      306.473 ms (23.65% GC)
-#   mean time:        330.366 ms (24.22% GC)
-#   maximum time:     487.338 ms (28.60% GC)
+#   minimum time:     127.855 ms (21.46% GC)
+#   median time:      169.408 ms (20.37% GC)
+#   mean time:        167.529 ms (20.88% GC)
+#   maximum time:     213.863 ms (21.28% GC)
 #   --------------
-#   samples:          16
+#   samples:          30
 #   evals/sample:     1
 
 bench_local_level = @benchmark statespace($locallevel, filter_type = $KalmanFilter, verbose = $verbose)
-# 1st August 2019
+# 5th August 2019
 # BenchmarkTools.Trial:
-#   memory estimate:  135.35 MiB
-#   allocs estimate:  1956723
+#   memory estimate:  128.48 MiB
+#   allocs estimate:  1832786
 #   --------------
-#   minimum time:     91.660 ms (24.23% GC)
-#   median time:      130.084 ms (25.73% GC)
-#   mean time:        128.891 ms (24.95% GC)
-#   maximum time:     172.001 ms (24.89% GC)
+#   minimum time:     88.344 ms (25.22% GC)
+#   median time:      132.428 ms (24.98% GC)
+#   mean time:        133.527 ms (24.29% GC)
+#   maximum time:     174.321 ms (23.58% GC)
 #   --------------
-#   samples:          39
+#   samples:          38
+#   evals/sample:     1
+
+bench_local_level = @benchmark statespace($locallevel, filter_type = $KalmanFilter, optimization_method = RandomSeedsLBFGS(; nseeds = 2), verbose = $verbose)
+# 5th August 2019
+# BenchmarkTools.Trial:
+#   memory estimate:  85.20 MiB
+#   allocs estimate:  1214590
+#   --------------
[…]

### 2019-08-05 src/kalman.jl

@@ -33,12 +33,10 @@ function kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}; t
     a[1, :]    = zeros(model.dim.m, 1)
     P[:, :, 1] = 1e6 .* Matrix(I, model.dim.m, model.dim.m)
 
-    missing_obs = check_missing_observation(model.y)
-
     RQR = model.R * Q * model.R'
     # Kalman filter
     for t = 1:model.dim.n
-        if t in missing_obs
+        if t in model.missing_observations
             steadystate  = false
             v[t, :]      = fill(NaN, model.dim.p)
             F[:, :, t]   = fill(NaN, (model.dim.p, model.dim.p))
@@ -105,10 +103,8 @@ function smoother(model::StateSpaceModel, kfilter::KalmanFilter)
     N[:, :, end] = zeros(m, m)
     r[end, :]    = zeros(m, 1)
 
-    missing_obs = check_missing_observation(v)
-
     @views @inbounds for t = n:-1:2
-        if t in missing_obs
+        if t in model.missing_observations
             r[t-1, :]     = T' * r[t, :]
             N[:, :, t-1]  = T' * N[:, :, t] * T
         else
@@ -122,7 +118,7 @@ function smoother(model::StateSpaceModel, kfilter::KalmanFilter)
     end
 
     # Last iteration
-    if 1 in missing_obs
+    if 1 in model.missing_observations
         r0  = T' * r[1, :]
         N0  = T' * N[:, :, 1] * T
     else

### 2019-08-05 src/sqrt_kalman.jl

@@ -42,7 +42,7 @@ function sqrt_kalman_filter(model::StateSpaceModel, sqrtH::Matrix{Typ}, sqrtQ::M
     sqrtH_zeros_pr  = [sqrtH zeros_pr]
     zeros_mp_RsqrtQ = [zeros_mp R*sqrtQ]
 
-    !isempty(check_missing_observation(model.y)) && error("Treatment of missing values not implemented for SquareRootFilter, please use KalmanFilter.")
+    !isempty(model.missing_observations) && error("Treatment of missing values not implemented for SquareRootFilter, please use KalmanFilter.")
 
     # Square-root Kalman filter
     for t = 1:n

### 2019-08-05 src/structures.jl

@@ -103,6 +103,7 @@ struct StateSpaceModel
     T::Matrix{Float64} # state matrix
     R::Matrix{Float64} # state error matrix
     dim::StateSpaceDimensions
+    missing_observations::Vector{Int}
     mode::String
 
     function StateSpaceModel(y::Matrix{Float64}, Z::Array{Float64, 3}, T::Matrix{Float64}, R::Matrix{Float64})
@@ -116,7 +117,7 @@ struct StateSpaceModel
             error("StateSpaceModel dimension mismatch")
         end
         dim = StateSpaceDimensions(ny, py, mr, rr)
-        new(y, Z, T, R, dim, "time-variant")
+        new(y, Z, T, R, dim, find_missing_observations(y), "time-variant")
     end
 
     function StateSpaceModel(y::Matrix{Float64}, Z::Matrix{Float64}, T::Matrix{Float64}, R::Matrix{Float64})
@@ -133,10 +134,10 @@ struct StateSpaceModel
 
         # Build Z
         Zvar = Array{Float64, 3}(undef, pz, mz, ny)
-        for t = 1:ny
-            Zvar[:, :, t] = Z
+        for t in 1:ny, i in axes(Z, 1), j in axes(Z, 2)
+            Zvar[i, j, t] = Z[i, j] # Zvar[:, :, t] = Z
         end
-        new(y, Zvar, T, R, dim, "time-invariant")
+        new(y, Zvar, T, R, dim, find_missing_observations(y), "time-invariant")
     end
 end
 

### 2019-08-05 src/utils.jl

@@ -85,7 +85,7 @@ function invertF(F::AbstractMatrix{T}) where T
     return size(F, 1) == 1 ? inv.(F) : inv(F)
 end
 
-function check_missing_observation(y::Matrix{T}) where T
+function find_missing_observations(y::Matrix{T}) where T
     missing_obs = Vector{Int}(undef, 0)
     for i in axes(y, 1), j in axes(y, 2)
         if isnan(y[i, j])

### 2019-08-05 src/kalman.jl
fix missing values performance regression (#85)
* fix missing values performance regression

* add benchmark and move missing_obsrvations inside the model

@@ -36,7 +36,7 @@ function kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}; t
     RQR = model.R * Q * model.R'
     # Kalman filter
     for t = 1:model.dim.n
-        if check_missing_observation(model.y, t)
+        if t in model.missing_observations
             steadystate  = false
             v[t, :]      = fill(NaN, model.dim.p)
             F[:, :, t]   = fill(NaN, (model.dim.p, model.dim.p))
@@ -104,7 +104,7 @@ function smoother(model::StateSpaceModel, kfilter::KalmanFilter)
     r[end, :]    = zeros(m, 1)
 
     @views @inbounds for t = n:-1:2
-        if check_missing_observation(v, t)
+        if t in model.missing_observations
             r[t-1, :]     = T' * r[t, :]
             N[:, :, t-1]  = T' * N[:, :, t] * T
         else
@@ -118,7 +118,7 @@ function smoother(model::StateSpaceModel, kfilter::KalmanFilter)
     end
 
     # Last iteration
-    if check_missing_observation(v, 1)
+    if 1 in model.missing_observations
         r0  = T' * r[1, :]
         N0  = T' * N[:, :, 1] * T
     else

### 2019-08-05 src/sqrt_kalman.jl

@@ -42,9 +42,10 @@ function sqrt_kalman_filter(model::StateSpaceModel, sqrtH::Matrix{Typ}, sqrtQ::M
     sqrtH_zeros_pr  = [sqrtH zeros_pr]
     zeros_mp_RsqrtQ = [zeros_mp R*sqrtQ]
 
+    !isempty(model.missing_observations) && error("Treatment of missing values not implemented for SquareRootFilter, please use KalmanFilter.")
+
     # Square-root Kalman filter
     for t = 1:n
-        any(isnan.(y[t, :])) && error("Treatment of missing values not implemented for SquareRootFilter, please use KalmanFilter.")
         v[t, :] = y[t, :] - Z[:, :, t]*a[t, :]
         if steadystate
             sqrtF[:, :, t] = sqrtF[:, :, t-1]

### 2019-08-05 src/utils.jl

@@ -85,8 +85,15 @@ function invertF(F::AbstractMatrix{T}) where T
     return size(F, 1) == 1 ? inv.(F) : inv(F)
 end
 
-function check_missing_observation(y::Matrix{T}, t::Int) where T
-    return any(isnan, view(y, t, :))
+function find_missing_observations(y::Matrix{T}) where T
+    missing_obs = Vector{Int}(undef, 0)
+    for i in axes(y, 1), j in axes(y, 2)
+        if isnan(y[i, j])
+            # If is empty or if the index i is not already pushed
+            (isempty(missing_obs) || (missing_obs[end] != i)) && push!(missing_obs, i)
+        end
+    end
+    return missing_obs
 end
 
 function Base.show(io::IO, ss::StateSpace)

### 2019-08-05 src/sqrt_kalman.jl
fix check steady state for sqrt filter (#86)

@@ -5,6 +5,8 @@ Square-root Kalman filter with big Kappa initialization.
 """
 function sqrt_kalman_filter(model::StateSpaceModel, sqrtH::Matrix{Typ}, sqrtQ::Matrix{Typ}; tol::Typ = 1e-5) where Typ <: AbstractFloat
 
+    time_invariant = model.mode == "time-invariant"
+
     # Load dimensions
     n, p, m, r = size(model)
 
@@ -67,7 +69,7 @@ function sqrt_kalman_filter(model::StateSpaceModel, sqrtH::Matrix{Typ}, sqrtQ::M
             sqrtP[:, :, t+1] = Ustar[range1, range1]
 
             # Checking if steady state was attained
-            if check_steady_state(sqrtP[:, :, t+1], sqrtP[:, :, t], tol)
+            if time_invariant && check_steady_state(sqrtP, t, tol)
                 steadystate = true
                 tsteady     = t
             end

### 2019-08-05 src/kalman_utils.jl
minor performance update

@@ -1,4 +1,20 @@
 # Utils for filter performance
+function fill_a1(a::Matrix{T}) where T <: AbstractFloat
+    for i in axes(a, 2)
+        a[1, i] = zero(T)
+    end
+end
+
+function fill_P1(P::AbstractArray{T}; bigkappa::Float64 = 1e6) where T <: AbstractFloat
+    for i in axes(P, 1), j in axes(P, 2)
+        if i == j
+            P[i, j, 1] = bigkappa
+        else
+            P[i, j, 1] = zero(T)
+        end
+    end
+end
+
 function repeat_matrix_t_plus_1(mat::AbstractArray{T}, t::Int) where T <: AbstractFloat
     @inbounds for j in axes(mat, 2), i in axes(mat, 1)
           mat[i, j, t+1] = mat[i, j, t]

### 2019-08-06 benchmark/bench.jl
minor performance update (#87)

@@ -32,16 +32,18 @@ bench_local_level = @benchmark statespace($locallevel, filter_type = $KalmanFilt
 #   samples:          38
 #   evals/sample:     1
 
-bench_local_level = @benchmark statespace($locallevel, filter_type = $KalmanFilter, optimization_method = RandomSeedsLBFGS(; nseeds = 2), verbose = $verbose)
+H = fill(1.0, (1,1))
+Q = fill(1.0, (1,1))
+@benchmark StateSpaceModels.kalman_filter($locallevel, $H, $Q)
 # 5th August 2019
 # BenchmarkTools.Trial:
-#   memory estimate:  85.20 MiB
-#   allocs estimate:  1214590
+#   memory estimate:  128.42 KiB
+#   allocs estimate:  1761
 #   --------------
-#   minimum time:     59.862 ms (25.65% GC)
-#   median time:      93.516 ms (23.43% GC)
-#   mean time:        93.485 ms (24.29% GC)
-#   maximum time:     137.639 ms (29.21% GC)
+#   minimum time:     63.112 μs (0.00% GC)
+#   median time:      69.907 μs (0.00% GC)
+#   mean time:        93.078 μs (21.86% GC)
+#   maximum time:     7.084 ms (98.82% GC)
 #   --------------
-#   samples:          54
+#   samples:          10000
 #   evals/sample:     1

### 2019-08-06 src/kalman.jl

@@ -30,10 +30,10 @@ function kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}; t
     tsteady     = model.dim.n+1
 
     # Initial state: big Kappa initialization
-    a[1, :]    = zeros(model.dim.m, 1)
-    P[:, :, 1] = 1e6 .* Matrix(I, model.dim.m, model.dim.m)
+    fill_a1(a)
+    fill_P1(P; bigkappa = 1e6)
 
-    RQR = model.R * Q * model.R'
+    RQR = model.R * LinearAlgebra.BLAS.gemm('N', 'T', 1.0, Q, model.R) # RQR = R Q R'
     # Kalman filter
     for t = 1:model.dim.n
         if t in model.missing_observations

### 2019-08-06 src/structures.jl
minor doc improvements

@@ -74,7 +74,7 @@ StateSpaceModel dimensions, following the notation of on the book
 * `n` is the number of observations
 * `p` is the dimension of the observation vector ``y_t``
 * `m` is the dimension of the state vector ``\\alpha_t``
-* `r` is the dimension of the state covariance matrix ``Q_t``
+* `r` is the dimension of the state covariance matrix ``Q``
 """
 struct StateSpaceDimensions
     n::Int
@@ -146,8 +146,8 @@ end
 
 Following the notation of on the book \"Time Series Analysis by State Space Methods\" (2012) by J. Durbin and S. J. Koopman.
 
-* `H` covariance matrix of the observation vector ``H_t``
-* `Q` covariance matrix of the state vector ``Q_t``
+* `H` covariance matrix of the observation vector
+* `Q` covariance matrix of the state vector
 """
 struct StateSpaceCovariance
     H::Matrix{Float64}
@@ -183,7 +183,7 @@ end
 Following the notation of on the book \"Time Series Analysis by State Space Methods\" (2012) by J. Durbin and S. J. Koopman.
 
 * `a` predictive states ``E(\\alpha_t|y_{t-1}, \\dots , y_1)``
-* `v` Prediction errors ``v_t = y_t − Z_ta_t,i, i = 1, \\dots , p``
+* `v` Prediction errors ``v_t = y_t − Z_t a_t``
 * `P` Covariance matrix of predictive states ``P = Var(\\alpha_t|y_{t−1}, \\dots , y_1)``
 * `F` Prediction error variances ``Var(v_t)``
 * `steadystate` Boolean to indicate if steady state was attained
@@ -203,7 +203,7 @@ end
 """
     StateSpace
 
-A state space structure containing the model, filter output, smoother output, covariance matrices, filter type and optimization method.
+A state-space structure containing the model, filter output, smoother output, covariance matrices, filter type and optimization method.
 """
 struct StateSpace
     model::StateSpaceModel

### 2019-08-06 src/kalman_utils.jl
update_v and update_att

@@ -49,12 +49,24 @@ function update_P(P::AbstractArray{Typ}, T::AbstractArray{Typ}, Ptt::AbstractArr
 end
 
 function update_v(v::AbstractArray{T}, y::AbstractArray{T}, Z::AbstractArray{T}, a::AbstractArray{T}, t::Int) where T <: AbstractFloat
-    @views @inbounds v[t, :] = y[t, :] - Z[:, :, t] * a[t, :]
+    # v[t, :] = y[t, :] - Z[:, :, t]*a[t, :]
+    for i in axes(Z, 1)
+        v[t, i] = y[t, i]
+        for j in axes(Z, 2)
+            v[t, i] -= Z[i, j, t]*a[t, j]
+        end
+    end
     return 
 end
 
 function update_att(att::AbstractArray{T}, a::AbstractArray{T}, P_Ztransp_invF::AbstractArray{T}, v::AbstractArray{T}, t::Int) where T <: AbstractFloat
-    @inbounds @views att[t, :] = a[t, :] + P_Ztransp_invF * v[t, :]
+    # att[t, :] = a[t, :] + P_Ztransp_invF*v[t, :]
+    for i in axes(P_Ztransp_invF, 1)
+        att[t, i] = a[t, i]
+        for j in axes(P_Ztransp_invF, 2)
+            att[t, i] += P_Ztransp_invF[i,j]*v[t, j]
+        end
+    end
     return 
 end
 

### 2019-08-06 src/structures.jl
adding filtered state to docstrings

@@ -146,8 +146,8 @@ end
 
 Following the notation of on the book \"Time Series Analysis by State Space Methods\" (2012) by J. Durbin and S. J. Koopman.
 
-* `H` covariance matrix of the observation vector
-* `Q` covariance matrix of the state vector
+* `H` Covariance matrix of the observation vector
+* `Q` Covariance matrix of the state vector
 """
 struct StateSpaceCovariance
     H::Matrix{Float64}
@@ -182,12 +182,14 @@ end
 
 Following the notation of on the book \"Time Series Analysis by State Space Methods\" (2012) by J. Durbin and S. J. Koopman.
 
-* `a` predictive states ``E(\\alpha_t|y_{t-1}, \\dots , y_1)``
-* `v` Prediction errors ``v_t = y_t − Z_t a_t``
-* `P` Covariance matrix of predictive states ``P = Var(\\alpha_t|y_{t−1}, \\dots , y_1)``
-* `F` Prediction error variances ``Var(v_t)``
+* `a` Predictive state ``E(\\alpha_t|y_{t-1}, \\dots , y_1)``
+* `att` Filtered state ``E(\\alpha_t|y_{t}, \\dots , y_1)``
+* `v` Prediction error ``v_t = y_t − Z_t a_t``
+* `P` Covariance matrix of predictive state ``P = Var(\\alpha_t|y_{t−1}, \\dots , y_1)``
+* `Ptt` Covariance matrix of filtered state ``P = Var(\\alpha_t|y_{t}, \\dots , y_1)``
+* `F` Variance of prediction error ``Var(v_{t})``
 * `steadystate` Boolean to indicate if steady state was attained
-* `tsteady` Instant when steady state was attained; in case it wasn't, `tsteady = n+1`
+* `tsteady` Instant when steady state was attained; in case it was not, `tsteady = n+1`
 """
 struct FilterOutput
     a::Matrix{Float64} # predictive state

### 2019-08-06 benchmark/bench.jl
update_v and update_att (#89)

@@ -19,17 +19,17 @@ bench_local_level = @benchmark statespace($locallevel, filter_type = $SquareRoot
 #   evals/sample:     1
 
 bench_local_level = @benchmark statespace($locallevel, filter_type = $KalmanFilter, verbose = $verbose)
-# 5th August 2019
+# 6th August 2019
 # BenchmarkTools.Trial:
-#   memory estimate:  128.48 MiB
-#   allocs estimate:  1832786
+#   memory estimate:  106.18 MiB
+#   allocs estimate:  1500659
 #   --------------
-#   minimum time:     88.344 ms (25.22% GC)
-#   median time:      132.428 ms (24.98% GC)
-#   mean time:        133.527 ms (24.29% GC)
-#   maximum time:     174.321 ms (23.58% GC)
+#   minimum time:     80.464 ms (22.75% GC)
+#   median time:      110.805 ms (23.33% GC)
+#   mean time:        112.070 ms (22.92% GC)
+#   maximum time:     131.950 ms (21.28% GC)
 #   --------------
-#   samples:          38
+#   samples:          45
 #   evals/sample:     1
 
 H = fill(1.0, (1,1))
@@ -37,13 +37,13 @@ Q = fill(1.0, (1,1))
 @benchmark StateSpaceModels.kalman_filter($locallevel, $H, $Q)
 # 5th August 2019
 # BenchmarkTools.Trial:
-#   memory estimate:  128.42 KiB
-#   allocs estimate:  1761
+#   memory estimate:  91.80 KiB
+#   allocs estimate:  1229
 #   --------------
-#   minimum time:     63.112 μs (0.00% GC)
-#   median time:      69.907 μs (0.00% GC)
-#   mean time:        93.078 μs (21.86% GC)
-#   maximum time:     7.084 ms (98.82% GC)
+#   minimum time:     49.136 μs (0.00% GC)
+#   median time:      56.570 μs (0.00% GC)
+#   mean time:        73.994 μs (20.43% GC)
+#   maximum time:     6.436 ms (97.36% GC)
 #   --------------
 #   samples:          10000
 #   evals/sample:     1

### 2019-10-18 paper/paper.tex
Correct typo
updating paper

@@
 Section \ref{sec:estimation} explains the filtering, estimation and smoothing processes.
+Section \ref{sec:diagnostics} discusses the available diagnostics to evaluate the model specification and estimation.
 In Section \ref{sec:forecasting}, the forecasting and simulation procedures are presented.
 Section \ref{sec:applications} contains examples of applications that illustrate the use of the package.
-Finally, conclusions and future work are provided in Section \ref{sec:conclusion}
+Finally, conclusions and future work are provided in Section \ref{sec:conclusion}.
 
@@
 \begin{lstlisting}[language = Julia]
-model = StateSpaceModel(y, Z, T, R)
+model = StateSpaceModel{Float64}(y, Z, T, R)
 \end{lstlisting}
@@
 Conversely, if $Z_{t}$ is time-varying, it must be input as an \texttt{Array\{Float64, 3\}} where the third dimension represents the time periods.
+Any desired float precision can be employed within the model.
 
@@
 \begin{lstlisting}[language = Julia]
-ss = statespace(model; filter_type = SquareRootFilter)
+ss = statespace(model; filter_type = SquareRootFilter{Float64})
 \end{lstlisting}
 
-If no \texttt{filter\_type} is provided, the standard type \texttt{KalmanFilter} is utilized as a default option.
+Filters allow any desired float precision.
+If no \texttt{filter\_type} is provided, the standard type \texttt{KalmanFilter\{Float64\}} is utilized as a default option.
 Further information on Kalman filtering can be found in Chapter 4.3 of \cite{durbin2012time}.
@@
 
+\section{Diagnostics} \label{sec:diagnostics}
+
+As a way of evaluating the model specification and estimation, the package contains statistical diagnostics that can be conducted over the residuals.
+The available diagnostics are the Jarque-Bera normality test \cite{jarque1980efficient}, the Ljung-Box independence test \cite{box1970distribution}, and an homoscedasticity test \cite{durbin2012time}.
+The diagnostics can be run through function \texttt{diagnostics}.
+
+An example of the diagnostics log can be seen below.
+%
+\begin{lstlisting}[language = Julia]
+julia> diagnostics(ss)
+==============================================================
+                    Running diagnostics...
+                Jarque-Bera: p-value = 0.89683
+                  Ljung-Box: p-value = 0.95855
+           Homoscedasticity: p-value = 0.30752
+==============================================================
+\end{lstlisting}
+
 \section{Forecasting and Simulation} \label{sec:forecasting}
@@
 # Estimate the state-space model
-ss = statespace(model; filter_type = SquareRootFilter)
+ss = statespace(model; filter_type = SquareRootFilter{Float64})
 # Number of months ahead to be simulated
@@
 # User defined model
-model = StateSpaceModel(y, Z, T, R)
+model = StateSpaceModel{Float64}(y, Z, T, R)
 # Estimate vehicle speed and position
@@
 
-In this example, we define the noise variances $H$ and $Q$, generate the noises and simulate a random vehicle trajectory using the state-space equations:
+In this example, we define the noise variances $H$ and $Q$, generate the noises and simulate a random vehicle trajectory using the state-space equations.
 %

### 2019-10-21 src/StateSpaceModels.jl
change AbstractFloat to Real, fix some docstrings

@@ -25,7 +25,7 @@ include("diagnostics.jl")
 Estimate the pre-specified state-space model.
 """
 function statespace(model::StateSpaceModel{T}; filter_type::DataType = KalmanFilter{T}, 
-                    optimization_method::AbstractOptimizationMethod = RandomSeedsLBFGS(), verbose::Int = 1) where T <: AbstractFloat
+                    optimization_method::AbstractOptimizationMethod = RandomSeedsLBFGS(), verbose::Int = 1) where T
 
     if !(verbose in [0, 1, 2])
         @warn("Incorrect verbose value input (should be 0, 1, or 2): switching to default value 1")
@@ -49,10 +49,10 @@ end
     kfas(model::StateSpaceModel, covariance::StateSpaceCovariance,
          filter_type::DataType)
 
-Perform kalman filter and smoother according to the chosen `filter_type`.
+Perform Kalman filter and smoother according to the chosen `filter_type`.
 """
 function kfas(model::StateSpaceModel{T}, covariance::StateSpaceCovariance{T},
-              filter_type::DataType) where T <: AbstractFloat
+              filter_type::DataType) where T
     error(filter_type , " not implemented") # Returns an error if it cannot
                                             # find a specialized kfas
 end

### 2019-10-21 src/forecast.jl

@@ -1,12 +1,12 @@
 export forecast, simulate
 
 """
-    forecast(ss::StateSpace, N::Int)
+    forecast(ss::StateSpace{Typ}, N::Int) where Typ
 
 Obtain the minimum mean square error forecasts N steps ahead. Returns the forecasts and the predictive distributions 
 at each time period.
 """
-function forecast(ss::StateSpace{Typ}, N::Int) where Typ <: AbstractFloat
+function forecast(ss::StateSpace{Typ}, N::Int) where Typ
 
     # Load estimated covariance matrices
     H = ss.covariance.H
@@ -53,12 +53,12 @@ function forecast(ss::StateSpace{Typ}, N::Int) where Typ <: AbstractFloat
 end
 
 """
-    simulate(ss::StateSpace, N::Int, S::Int)
+    simulate(ss::StateSpace{Typ}, N::Int, S::Int) where Typ
 
 Simulate S future scenarios up to N steps ahead. Returns a p x N x S matrix where the dimensions represent, respectively,
 the number of series in the model, the number of steps ahead, and the number of scenarios.
 """
-function simulate(ss::StateSpace{Typ}, N::Int, S::Int) where Typ <: AbstractFloat
+function simulate(ss::StateSpace{Typ}, N::Int, S::Int) where Typ
 
     # Load estimated covariance matrices
     H = ss.covariance.H
@@ -96,11 +96,11 @@ function simulate(ss::StateSpace{Typ}, N::Int, S::Int) where Typ <: AbstractFloa
 end
 
 """
-    prepare_forecast(ss::StateSpace, N::Int)
+    prepare_forecast(ss::StateSpace{Typ}, N::Int) where Typ
 
 Adjust matrix Z for forecasting and check for dimension errors.
 """
-function prepare_forecast(ss::StateSpace{Typ}, N::Int) where Typ <: AbstractFloat
+function prepare_forecast(ss::StateSpace{Typ}, N::Int) where Typ
 
     # Load system
     n, p, m, r = size(ss.model)

### 2019-10-21 src/kalman_utils.jl

@@ -1,10 +1,10 @@
 # Utils for filter performance
-function fill_a1!(a::Matrix{T}) where T <: AbstractFloat
+function fill_a1!(a::Matrix{T}) where T
     @inbounds for i in axes(a, 2)
         a[1, i] = zero(T)
     end
 end
-function fill_P1!(P::AbstractArray{T}; bigkappa::T = 1e6) where T <: AbstractFloat
+function fill_P1!(P::AbstractArray{T}; bigkappa::T = 1e6) where T
     @inbounds for i in axes(P, 1), j in axes(P, 2)
         if i == j
             P[i, j, 1] = bigkappa
@@ -14,24 +14,24 @@ function fill_P1!(P::AbstractArray{T}; bigkappa::T = 1e6) where T <: AbstractFlo
     end
 end
 
-function repeat_matrix_t_plus_1!(mat::AbstractArray{T}, t::Int) where T <: AbstractFloat
+function repeat_matrix_t_plus_1!(mat::AbstractArray{T}, t::Int) where T
     @inbounds for j in axes(mat, 2), i in axes(mat, 1)
           mat[i, j, t+1] = mat[i, j, t]
     end
     return 
 end
-function repeat_vector_t_plus_1!(mat::Matrix{T}, t::Int) where T <: AbstractFloat
+function repeat_vector_t_plus_1!(mat::Matrix{T}, t::Int) where T
     @inbounds for i in axes(mat, 1)
         mat[i, t+1] = mat[i, t]
     end
     return 
 end
 
-function big_update_a!(a::Matrix{Typ}, att::Matrix{Typ}, T::Matrix{Typ}, t::Int) where Typ <: AbstractFloat
+function big_update_a!(a::Matrix{Typ}, att::Matrix{Typ}, T::Matrix{Typ}, t::Int) where Typ
     @views @inbounds mul!(a[t+1, :], T, att[t, :])
     return 
 end
-function small_update_a!(a::Matrix{Typ}, att::Matrix{Typ}, T::Matrix{Typ}, t::Int) where Typ <: AbstractFloat
+function small_update_a!(a::Matrix{Typ}, att::Matrix{Typ}, T::Matrix{Typ}, t::Int) where Typ
     @inbounds for i in axes(a, 2)
         a[t+1, i] = zero(Typ)
[…]

### 2019-10-21 src/sqrt_kalman.jl

@@ -1,9 +1,9 @@
 """
-    sqrt_kalman_filter(model::StateSpaceModel, sqrtH::Matrix{Typ}, sqrtQ::Matrix{Typ}; tol::Typ = 1e-5) where Typ <: AbstractFloat
+    sqrt_kalman_filter(model::StateSpaceModel{Typ}, sqrtH::Matrix{Typ}, sqrtQ::Matrix{Typ}; tol::Typ = 1e-5) where Typ
 
 Square-root Kalman filter with big Kappa initialization.
 """
-function sqrt_kalman_filter(model::StateSpaceModel{Typ}, sqrtH::Matrix{Typ}, sqrtQ::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ <: AbstractFloat
+function sqrt_kalman_filter(model::StateSpaceModel{Typ}, sqrtH::Matrix{Typ}, sqrtQ::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
     time_invariant = model.mode == "time-invariant"
 
@@ -81,11 +81,11 @@ function sqrt_kalman_filter(model::StateSpaceModel{Typ}, sqrtH::Matrix{Typ}, sqr
 end
 
 """
-    filtered_state(model::StateSpaceModel, sqrt_filter::SquareRootFilter)
+    filtered_state(model::StateSpaceModel{Typ}, sqrt_filter::SquareRootFilter{Typ}) where Typ
 
 Obtain the filtered state estimates and their covariance matrices.
 """
-function filtered_state(model::StateSpaceModel{Typ}, sqrt_filter::SquareRootFilter{Typ}) where Typ <: AbstractFloat
+function filtered_state(model::StateSpaceModel{Typ}, sqrt_filter::SquareRootFilter{Typ}) where Typ
 
     # Load dimensions data
     n, p, m, r = size(model)
@@ -115,11 +115,11 @@ function filtered_state(model::StateSpaceModel{Typ}, sqrt_filter::SquareRootFilt
 end
 
 """
-    sqrt_smoother(model::StateSpaceModel, sqrt_filter::SquareRootFilter)
+    sqrt_smoother(model::StateSpaceModel{Typ}, sqrt_filter::SquareRootFilter{Typ}) where Typ
 
 Square-root smoother for state space model.
 """
[…]

### 2019-10-21 src/univariate_kalman.jl

@@ -1,9 +1,9 @@
 """
-    kalman_filter(model::StateSpaceModel, H::Typ, Q::Matrix{Typ}; tol::Typ = 1e-5) where Typ <: AbstractFloat
+    univariate_kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
 Kalman filter with big Kappa initialization, i.e., initializing state variances as 1e6.
 """
-function univariate_kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ <: AbstractFloat
+function univariate_kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
     time_invariant = model.mode == "time-invariant"
 
@@ -73,11 +73,11 @@ function univariate_kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Mat
 end
 
 """
-    univariate_smoother(model::StateSpaceModel, kfilter::UnivariateKalmanFilter)
+    univariate_smoother(model::StateSpaceModel{Typ}, kfilter::UnivariateKalmanFilter{Typ}) where Typ
 
 Smoother for state-space model.
 """
-function univariate_smoother(model::StateSpaceModel{Typ}, kfilter::UnivariateKalmanFilter{Typ}) where Typ <: AbstractFloat
+function univariate_smoother(model::StateSpaceModel{Typ}, kfilter::UnivariateKalmanFilter{Typ}) where Typ
 
     # Load dimensions data
     n, p, m, r = size(model)
@@ -139,7 +139,7 @@ end
 # *
 
 function statespace_covariance(psi::Vector{T}, p::Int, r::Int,
-                               filter_type::Type{UnivariateKalmanFilter{T}}) where T <: AbstractFloat
+                               filter_type::Type{UnivariateKalmanFilter{T}}) where T
 
     # Build lower triangular matrices
     H = ones(1, 1)*psi[1]^2
@@ -155,8 +155,7 @@ function statespace_covariance(psi::Vector{T}, p::Int, r::Int,
 end
 
 function get_log_likelihood_params(psitilde::Vector{T}, model::StateSpaceModel,
-                                   filter_type::Type{UnivariateKalmanFilter{T}}) where T <: AbstractFloat
-
[…]

### 2019-10-21 src/diagnostics.jl
change AbstractFloat to Real, fix some docstrings (#107)

@@ -33,7 +33,7 @@ end
 
 Obtain standardized residuals from a state-space model
 """
-function residuals(ss::StateSpace{T}) where T <: AbstractFloat
+function residuals(ss::StateSpace{T}) where T
     e = Matrix{T}(undef, ss.model.dim.n, ss.model.dim.p)
     for j = 1:ss.model.dim.p
         for t = 1:ss.model.dim.n
@@ -52,7 +52,7 @@ end
 
 Run Jarque-Bera normality test and return p-values
 """
-function jarquebera(e::Matrix{T}) where T <: AbstractFloat
+function jarquebera(e::Matrix{T}) where T
     n, p = size(e)
     pvalue = Vector{T}(undef, p)
     for j = 1:p
@@ -74,7 +74,7 @@ end
 
 Run Ljung-Box independence test and return p-values
 """
-function ljungbox(e::Matrix{T}; maxlag::Int = 20) where T <: AbstractFloat
+function ljungbox(e::Matrix{T}; maxlag::Int = 20) where T
     n, p = size(e)
     n <= maxlag && (maxlag = n-1)
     pvalue = Vector{T}(undef, p)
@@ -93,7 +93,7 @@ end
 
 Run homoscedasticity test and return p-values
 """
-function homoscedast(e::Matrix{T}) where T <: AbstractFloat
+function homoscedast(e::Matrix{T}) where T
     n, p = size(e)
     t1 = floor(Int, n/3)
     t2 = ceil(Int, 2*(n/3)+1)

### 2019-10-21 src/estimation.jl

@@ -1,11 +1,12 @@
-function compute_log_likelihood(n::Int, p::Int, v::Matrix{T}, F::Array{T, 3}, valid_insts::Vector{Int}) where T <: AbstractFloat
+function compute_log_likelihood(n::Int, p::Int, v::Matrix{T}, F::Array{T, 3}, valid_insts::Vector{Int}) where T
     if p == 1
         return compute_log_likelihood_univariate(n, v, F, valid_insts)
     else
         return compute_log_likelihood_multivariate(n, p, v, F, valid_insts)
     end
 end
-function compute_log_likelihood(n::Int, p::Int, v::Vector{T}, F::Vector{T}, valid_insts::Vector{Int}) where T <: AbstractFloat
+
+function compute_log_likelihood(n::Int, p::Int, v::Vector{T}, F::Vector{T}, valid_insts::Vector{Int}) where T
     log_likelihood::T = n*log(2*pi)/2
     @inbounds for t in valid_insts
         log_likelihood += 0.5 * (log(F[t]) + (v[t]^2)/F[t])
@@ -13,7 +14,7 @@ function compute_log_likelihood(n::Int, p::Int, v::Vector{T}, F::Vector{T}, vali
     return log_likelihood
 end
 
-function compute_log_likelihood_univariate(n::Int, v::Matrix{T}, F::Array{T, 3}, valid_insts::Vector{Int}) where T <: AbstractFloat
+function compute_log_likelihood_univariate(n::Int, v::Matrix{T}, F::Array{T, 3}, valid_insts::Vector{Int}) where T
     log_likelihood::T = n*log(2*pi)/2
     @inbounds for t in valid_insts
         log_likelihood += 0.5 * (log(F[1, 1, t]) + (v[t, 1]^2)/F[1, 1, t])
@@ -21,7 +22,7 @@ function compute_log_likelihood_univariate(n::Int, v::Matrix{T}, F::Array{T, 3},
     return log_likelihood
 end
 
-function compute_log_likelihood_multivariate(n::Int, p::Int, v::Matrix{T}, F::Array{T, 3}, valid_insts::Vector{Int}) where T <: AbstractFloat
+function compute_log_likelihood_multivariate(n::Int, p::Int, v::Matrix{T}, F::Array{T, 3}, valid_insts::Vector{Int}) where T 
     log_likelihood::T = n*p*log(2*pi)/2
     @inbounds @views for t in valid_insts
[…]

### 2019-10-21 src/kalman.jl

@@ -1,9 +1,9 @@
 """
-    kalman_filter(model::StateSpaceModel, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = 1e-5) where Typ <: AbstractFloat
+    kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = 1e-5) where Typ <: Real
 
 Kalman filter with big Kappa initialization, i.e., initializing state variances as 1e6.
 """
-function kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ <: AbstractFloat
+function kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
     time_invariant = model.mode == "time-invariant"
 
@@ -73,11 +73,11 @@ function kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Ty
 end
 
 """
-    smoother(model::StateSpaceModel, kfilter::KalmanFilter)
+    smoother(model::StateSpaceModel{Typ}, kfilter::KalmanFilter{Typ}) where Typ
 
 Smoother for state-space model.
 """
-function smoother(model::StateSpaceModel{Typ}, kfilter::KalmanFilter{Typ}) where Typ <: AbstractFloat
+function smoother(model::StateSpaceModel{Typ}, kfilter::KalmanFilter{Typ}) where Typ
 
     # Load dimensions data
     n, p, m, r = size(model)
@@ -139,7 +139,7 @@ end
 # *
 
 function statespace_covariance(psi::Vector{T}, p::Int, r::Int,
-                               filter_type::Type{KalmanFilter{T}}) where T <: AbstractFloat
+                               filter_type::Type{KalmanFilter{T}}) where T
 
     # Build lower triangular matrices
     if p > 1
@@ -162,7 +162,7 @@ function statespace_covariance(psi::Vector{T}, p::Int, r::Int,
 end
 
 function get_log_likelihood_params(psitilde::Vector{T}, model::StateSpaceModel{T},
-                                   filter_type::Type{KalmanFilter{T}}) where T <: AbstractFloat
+                                   filter_type::Type{KalmanFilter{T}}) where T
 
[…]

### 2019-10-21 src/models.jl

@@ -1,14 +1,14 @@
 export structural, local_level, linear_trend
 
 """
-    structural(y::VecOrMat{Typ}, s::Int; X::VecOrMat{Typ} = Matrix{Typ}(undef, 0, 0)) where Typ <: AbstractFloat
+    structural(y::VecOrMat{Typ}, s::Int; X::VecOrMat{Typ} = Matrix{Typ}(undef, 0, 0)) where Typ <: Real
 
 Build state-space system for a given structural model with observations y, seasonality s, and, optionally, exogenous variables X.
 
 If `y` is provided as an `Array{Typ, 1}` it will be converted to an `Array{Typ, 2}` inside the `StateSpaceModel`. The same will happen to X, 
 if an `Array{Typ, 1}` it will be converted to an `Array{Typ, 2}` inside the `StateSpaceModel`.
 """
-function structural(y::VecOrMat{Typ}, s::Int; X::VecOrMat{Typ} = Matrix{Typ}(undef, 0, 0)) where Typ <: AbstractFloat
+function structural(y::VecOrMat{Typ}, s::Int; X::VecOrMat{Typ} = Matrix{Typ}(undef, 0, 0)) where Typ <: Real
 
     # Number of observations and endogenous variables
     y = y[:, :]
@@ -85,13 +85,13 @@ function structural(y::VecOrMat{Typ}, s::Int; X::VecOrMat{Typ} = Matrix{Typ}(und
 end
 
 """
-    local_level(y::VecOrMat{Typ}) where Typ <: AbstractFloat
+    local_level(y::VecOrMat{Typ}) where Typ <: Real
 
 Build state-space system for a local level model with observations y.
 
 If `y` is proided as an `Array{Typ, 1}` it will be converted to an `Array{Typ, 2}` inside the `StateSpaceModel`.
 """
-function local_level(y::VecOrMat{Typ}) where Typ <: AbstractFloat
+function local_level(y::VecOrMat{Typ}) where Typ <: Real
 
     # Number of observations and endogenous variables
     y = y[:, :]
@@ -111,13 +111,13 @@ function local_level(y::VecOrMat{Typ}) where Typ <: AbstractFloat
 end
 
 """
-    linear_trend(y::VecOrMat{Typ}) where Typ <: AbstractFloat
+    linear_trend(y::VecOrMat{Typ}) where Typ <: Real
 
 Build state-space system for a linear trend model with observations y.
 
[…]

### 2019-10-21 src/prints.jl

@@ -19,7 +19,7 @@ function print_estimation_start(verbose::Int, nseeds::Int)
     end
 end
 
-function print_loglikelihood(verbose::Int, iseed::Int, loglikelihood::Vector{T}, t0::DateTime) where T <: AbstractFloat
+function print_loglikelihood(verbose::Int, iseed::Int, loglikelihood::Vector{T}, t0::DateTime) where T
     if verbose > 0
         t1 = now() - t0
 
@@ -36,7 +36,7 @@ function print_loglikelihood(verbose::Int, iseed::Int, loglikelihood::Vector{T},
     end
 end
 
-function print_estimation_end(verbose::Int, loglikelihood::Vector{T}) where T <: AbstractFloat
+function print_estimation_end(verbose::Int, loglikelihood::Vector{T}) where T
     if verbose > 0
         str = "                  Log-likelihood: "
         str_loglik = @sprintf("%.4f", maximum(loglikelihood)) * "                  "

### 2019-10-21 src/random_seeds_lbfgs.jl

@@ -1,5 +1,5 @@
 function estimate_statespace(model::StateSpaceModel{T}, filter_type::DataType,
-                             opt_method::RandomSeedsLBFGS; verbose::Int = 1) where T <: AbstractFloat
+                             opt_method::RandomSeedsLBFGS; verbose::Int = 1) where T
 
     nseeds = opt_method.nseeds                         
     nseeds += 1 # creating additional seed for degenerate cases

### 2019-10-21 src/structures.jl

@@ -26,7 +26,7 @@ Abstract type used to implement an interface for generic optimization methods.
 abstract type AbstractOptimizationMethod end
 
 # Auxiliary structure for square-root Kalman filter
-mutable struct SquareRootFilter{T <: AbstractFloat} <: AbstractFilter
+mutable struct SquareRootFilter{T <: Real} <: AbstractFilter
     a::Matrix{T} # predictive state
     v::Matrix{T} # innovations
     sqrtP::Array{T, 3} # lower triangular matrix with sqrt-covariance of the predictive state
@@ -37,7 +37,7 @@ mutable struct SquareRootFilter{T <: AbstractFloat} <: AbstractFilter
 end
 
 # Auxiliary structure for Kalman filter
-mutable struct KalmanFilter{T <: AbstractFloat} <: AbstractFilter
+mutable struct KalmanFilter{T <: Real} <: AbstractFilter
     a::Matrix{T} # predictive state
     att::Matrix{T}
     v::Matrix{T} # innovations
@@ -50,7 +50,7 @@ mutable struct KalmanFilter{T <: AbstractFloat} <: AbstractFilter
 end
 
 # Auxiliary structure for univariate Kalman filter
-mutable struct UnivariateKalmanFilter{T <: AbstractFloat} <: AbstractFilter
+mutable struct UnivariateKalmanFilter{T <: Real} <: AbstractFilter
     a::Matrix{T} # predictive state
     att::Matrix{T}
     v::Vector{T} # innovations
@@ -63,24 +63,23 @@ mutable struct UnivariateKalmanFilter{T <: AbstractFloat} <: AbstractFilter
 end
 
 # Auxiliary structure for smoother
-mutable struct Smoother{T <: AbstractFloat} <: AbstractSmoother
+mutable struct Smoother{T <: Real} <: AbstractSmoother
     alpha::Matrix{T} # smoothed state
     V::Array{T, 3} # variance of smoothed state
 end
-
 
 # Concrete structs for optimization methods
 """
 TODO
 """
-mutable struct RandomSeedsLBFGS{T <: AbstractFloat} <: AbstractOptimizationMethod
+mutable struct RandomSeedsLBFGS{T <: Real} <: AbstractOptimizationMethod
     f_tol::T
     g_tol::T
     iterations::Int
     nseeds::Int
     seeds::Array{T}
 
[…]

### 2019-10-21 src/utils.jl

@@ -20,7 +20,7 @@ end
 
 
 # Linar Algebra wrappers
-function gram_in_time(mat::Array{T, 3}) where T <: AbstractFloat
+function gram_in_time(mat::Array{T, 3}) where T
     gram_in_time = similar(mat)
     @inbounds @views for t = 1:size(gram_in_time, 3)
         gram_in_time[:, :, t] = gram(mat[:, :, t])
@@ -28,7 +28,7 @@ function gram_in_time(mat::Array{T, 3}) where T <: AbstractFloat
     return gram_in_time
 end
 
-function gram(mat::AbstractArray{T}) where T <: AbstractFloat
+function gram(mat::AbstractArray{T}) where T
     if size(mat, 1) == 1
         gram_mat = Matrix{T}(undef, 1, 1)
         gram_mat[1, 1] = mat[1, 1]^2
@@ -39,16 +39,16 @@ function gram(mat::AbstractArray{T}) where T <: AbstractFloat
 end
 
 """
-    check_steady_state(P_t1::Matrix{T}, P_t::Matrix{T}, tol::T) where T <: AbstractFloat
+    check_steady_state(P_t1::Matrix{T}, P_t::Matrix{T}, tol::T) where T
 
 Return `true` if steady state was attained with respect to tolerance `tol`, `false` otherwise.
 The steady state is checked by the following equation. TODO
 """
-function check_steady_state(P_t1::Matrix{T}, P_t::Matrix{T}, tol::T) where T <: AbstractFloat
+function check_steady_state(P_t1::Matrix{T}, P_t::Matrix{T}, tol::T) where T
     return maximum(abs.((P_t1 - P_t)./P_t1)) < tol ? true : false
 end
 
-function check_steady_state(P::AbstractArray{T}, t::Int, tol::T) where T <: AbstractFloat
+function check_steady_state(P::AbstractArray{T}, t::Int, tol::T) where T
     @inbounds for j in axes(P, 2), i in axes(P, 1)
         if abs((P[i, j, t+1] - P[i, j, t])/P[i, j, t+1]) > tol
             return false
@@ -58,11 +58,11 @@ function check_steady_state(P::AbstractArray{T}, t::Int, tol::T) where T <: Abst
 end
 
 """
-    ensure_pos_sym!(M::Matrix{T}; ϵ::T = 1e-8) where T <: AbstractFloat
[…]

### 2019-10-21 src/models.jl
improving some docstrings

@@ -3,7 +3,7 @@ export structural, local_level, linear_trend
 """
     structural(y::VecOrMat{Typ}, s::Int; X::VecOrMat{Typ} = Matrix{Typ}(undef, 0, 0)) where Typ <: Real
 
-Build state-space system for a given structural model with observations y, seasonality s, and, optionally, exogenous variables X.
+Build state-space system for a given structural model with observations `y`, seasonality `s`, and, optionally, exogenous variables `X`.
 
 If `y` is provided as an `Array{Typ, 1}` it will be converted to an `Array{Typ, 2}` inside the `StateSpaceModel`. The same will happen to X, 
 if an `Array{Typ, 1}` it will be converted to an `Array{Typ, 2}` inside the `StateSpaceModel`.
@@ -89,7 +89,7 @@ end
 
 Build state-space system for a local level model with observations y.
 
-If `y` is proided as an `Array{Typ, 1}` it will be converted to an `Array{Typ, 2}` inside the `StateSpaceModel`.
+If `y` is provided as an `Array{Typ, 1}` it will be converted to an `Array{Typ, 2}` inside the `StateSpaceModel`.
 """
 function local_level(y::VecOrMat{Typ}) where Typ <: Real
 
@@ -115,7 +115,7 @@ end
 
 Build state-space system for a linear trend model with observations y.
 
-If `y` is proided as an `Array{Typ, 1}` it will be converted to an `Array{Typ, 2}` inside the `StateSpaceModel`.
+If `y` is provided as an `Array{Typ, 1}` it will be converted to an `Array{Typ, 2}` inside the `StateSpaceModel`.
 """
 function linear_trend(y::VecOrMat{Typ}) where Typ <: Real
 

### 2019-10-21 src/diagnostics.jl
improving some docstrings (#109)

@@ -1,9 +1,9 @@
 """
-    diagnostics(ss::StateSpace)
+    diagnostics(ss::StateSpace{T}) where T
 
 Run diagnostics and print results
 """
-function diagnostics(ss::StateSpace)
+function diagnostics(ss::StateSpace{T}) where T
     println("==============================================================")
     println("                    Running diagnostics...                    ")
     
@@ -29,7 +29,7 @@ function diagnostics(ss::StateSpace)
 end
 
 """
-    residuals(ss::StateSpace)
+    residuals(ss::StateSpace{T}) where T
 
 Obtain standardized residuals from a state-space model
 """
@@ -48,7 +48,7 @@ function residuals(ss::StateSpace{T}) where T
 end
 
 """
-    jarquebera(e::Matrix{T})
+    jarquebera(e::Matrix{T}) where T
 
 Run Jarque-Bera normality test and return p-values
 """
@@ -70,7 +70,7 @@ function jarquebera(e::Matrix{T}) where T
 end
 
 """
-    ljungbox(e::Matrix{T}; maxlag = 20)
+    ljungbox(e::Matrix{T}; maxlag::Int = 20) where T
 
 Run Ljung-Box independence test and return p-values
 """
@@ -89,7 +89,7 @@ function ljungbox(e::Matrix{T}; maxlag::Int = 20) where T
 end
 
 """
-    homoscedast(e::Matrix{T})
+    homoscedast(e::Matrix{T}) where T
 
 Run homoscedasticity test and return p-values
 """

### 2019-10-22 src/utils.jl
fixing multivariate and adding test

@@ -74,6 +74,18 @@ function ensure_pos_sym!(M::AbstractArray{T}, t::Int; ϵ::T = T(1e-8)) where T
     return 
 end
 
+function ensure_pos_sym!(M::AbstractArray{T}; ϵ::T = T(1e-8)) where T
+    @inbounds for j in axes(M, 2), i in 1:j
+        if i == j
+            M[i, i] = (M[i, i] + M[i, i])/2 + ϵ
+        else
+            M[i, j] = (M[i, j] + M[j, i])/2
+            M[j, i] = M[i, j]
+        end
+    end
+    return 
+end
+
 function sum_matrix!(mat_prin::AbstractArray{T}, mat_sum::AbstractMatrix{T}, t::Int, offset::Int) where T
     @inbounds for j in axes(mat_prin, 2), i in axes(mat_prin, 1)
         mat_prin[i, j, t + offset] = mat_prin[i, j, t + offset] + mat_sum[i, j]

### 2019-10-22 test/runtests.jl

@@ -1,4 +1,4 @@
-using Test, StateSpaceModels, Statistics, CSV, LinearAlgebra
+using Test, StateSpaceModels, Statistics, CSV, LinearAlgebra, Random
 
 const SSM = StateSpaceModels
 

### 2019-10-22 test/test_structural.jl

@@ -35,8 +35,7 @@
         compare_forecast_simulation(ss, 10, 1000, 1e-3)
     end
 
-    @testset "Multivariate test" begin
-       
+    @testset "Basic multivariate test" begin
         y = [ones(20) collect(1:20)]
         model = structural(y, 2)
 

### 2019-10-22 test/runtests.jl
adding distributions dependency on runtests

@@ -1,4 +1,4 @@
-using Test, StateSpaceModels, Statistics, CSV, LinearAlgebra, Random
+using Test, StateSpaceModels, Statistics, CSV, LinearAlgebra, Random, Distributions
 
 const SSM = StateSpaceModels
 

### 2019-10-22 test/test_locallevel.jl
added best seed optimization result

@@ -106,7 +106,7 @@ y = [0.0     # Deterministic local level generated time series
         @test isa(unimodel, StateSpaceModel)
         @test unimodel.mode == "time-invariant"
         
-        ss1 = statespace(unimodel)
+        ss1 = statespace(unimodel; verbose = 2)
         @test ss1.filter_type <: KalmanFilter
         @test isa(ss1, StateSpace)
         

### 2019-10-22 test/test_structural.jl
removing old multivariate test

@@ -35,24 +35,6 @@
         compare_forecast_simulation(ss, 10, 1000, 1e-3)
     end
 
-    @testset "Basic multivariate test" begin
-        y = [ones(20) collect(1:20)]
-        model = structural(y, 2)
-
-        @test isa(model, StateSpaceModel)
-        @test model.mode == "time-invariant"
-
-        ss = statespace(model)
-        diags = diagnostics(ss)
-        
-        sim  = simulate(ss, 10, 1000)
-
-        @test ss.filter_type <: KalmanFilter
-        @test mean(sim, dims = 3)[:, 1] ≈ ones(10) rtol = 1e-3
-        @test mean(sim, dims = 3)[:, 2] ≈ collect(21:30) rtol = 1e-3
-        compare_forecast_simulation(ss, 20, 1000, 1e-3)
-    end
-
     @testset "Error tests" begin
         y = ones(15, 1)
         @test_throws ErrorException structural(y, 2; X = ones(10, 2))

### 2019-10-23 src/univariate_kalman.jl
fixing multivariate and adding test (#110)
* fixing multivariate and adding test

* adding distributions dependency on runtests

* removing old multivariate test

@@ -145,7 +145,7 @@ function statespace_covariance(psi::Vector{T}, p::Int, r::Int,
     H = ones(1, 1)*psi[1]^2
     unknownsH = 1
 
-    sqrtQ = kron(Matrix{T}(I, Int(r/p), Int(r/p)), tril!(ones(p, p)))
+    sqrtQ = kron(tril!(ones(p, p)), Matrix{T}(I, Int(r/p), Int(r/p)))
     sqrtQ[findall(isequal(1), sqrtQ)] = psi[(unknownsH+1):Int(unknownsH + (r/p)*(p*(p + 1)/2))]
 
     # Obtain full matrices

### 2019-10-23 test/test_locallevel.jl

@@ -99,49 +99,78 @@ y = [0.0     # Deterministic local level generated time series
 6.165810835928076
 8.366964896750876]
 
-@testset "Local level" begin
-    @testset "Local level model with Kalman filter" begin
-        unimodel = local_level(y)
+@testset "Local level model with Kalman filter" begin
+    unimodel = local_level(y)
 
-        @test isa(unimodel, StateSpaceModel)
-        @test unimodel.mode == "time-invariant"
-        
-        ss1 = statespace(unimodel)
-        @test ss1.filter_type <: KalmanFilter
-        @test isa(ss1, StateSpace)
-        
-        ss2 = statespace(unimodel; filter_type = SquareRootFilter{Float64})
-        @test ss2.filter_type <: SquareRootFilter
-        @test isa(ss2, StateSpace)
-        
-        ss3 = statespace(unimodel; filter_type = UnivariateKalmanFilter{Float64})
-        @test ss3.filter_type <: UnivariateKalmanFilter
-        @test isa(ss3, StateSpace)
+    @test isa(unimodel, StateSpaceModel)
+    @test unimodel.mode == "time-invariant"
+    
+    ss1 = statespace(unimodel)
+    @test ss1.filter_type <: KalmanFilter
+    @test isa(ss1, StateSpace)
+    
+    ss2 = statespace(unimodel; filter_type = SquareRootFilter{Float64})
+    @test ss2.filter_type <: SquareRootFilter
+    @test isa(ss2, StateSpace)
+    
+    ss3 = statespace(unimodel; filter_type = UnivariateKalmanFilter{Float64})
+    @test ss3.filter_type <: UnivariateKalmanFilter
+    @test isa(ss3, StateSpace)
 
-        @test ss2.smoother.alpha ≈ ss1.smoother.alpha rtol = 1e-3
-        @test ss3.smoother.alpha ≈ ss1.smoother.alpha rtol = 1e-3
-        @test ss2.filter.a ≈ ss1.filter.a rtol = 1e-3
-        @test ss3.filter.a ≈ ss1.filter.a rtol = 1e-3
+    @test ss2.smoother.alpha ≈ ss1.smoother.alpha rtol = 1e-3
+    @test ss3.smoother.alpha ≈ ss1.smoother.alpha rtol = 1e-3
[…]

### 2019-10-23 test/test_structural.jl

@@ -35,25 +35,6 @@
         compare_forecast_simulation(ss, 10, 1000, 1e-3)
     end
 
-    @testset "Multivariate test" begin
-       
-        y = [ones(20) collect(1:20)]
-        model = structural(y, 2)
-
-        @test isa(model, StateSpaceModel)
-        @test model.mode == "time-invariant"
-
-        ss = statespace(model)
-        diags = diagnostics(ss)
-        
-        sim  = simulate(ss, 10, 1000)
-
-        @test ss.filter_type <: KalmanFilter
-        @test mean(sim, dims = 3)[:, 1] ≈ ones(10) rtol = 1e-3
-        @test mean(sim, dims = 3)[:, 2] ≈ collect(21:30) rtol = 1e-3
-        compare_forecast_simulation(ss, 20, 1000, 1e-3)
-    end
-
     @testset "Error tests" begin
         y = ones(15, 1)
         @test_throws ErrorException structural(y, 2; X = ones(10, 2))

### 2019-10-24 paper/paper.tex
Updating paper with new types and diagnostics (#105)
* Updating paper with new types and diagnostics
Add best seed optimization result as an option in verbose (#111)
* added best seed optimization result

* update paper.pdf and paper.tex

@@
 In this step, the Kalman filtering, the maximum likelihood estimation of the fixed parameters, and the smoothing are conducted.
-The optional argument \texttt{verbose}, which defaults to 1, specifies the verbosity: 0 for no output, 1 for some progress information, and 2 for progress information as well as the optimization log.
+The optional argument \texttt{verbose}, which defaults to 1, specifies the verbosity: 0 for no output, 1 for some progress information, 2 for progress information and summary of the optimization result and 3 for progress information as well as the full optimization log.
 

### 2019-10-24 src/StateSpaceModels.jl

@@ -27,8 +27,8 @@ Estimate the pre-specified state-space model.
 function statespace(model::StateSpaceModel{T}; filter_type::DataType = KalmanFilter{T}, 
                     optimization_method::AbstractOptimizationMethod = RandomSeedsLBFGS(), verbose::Int = 1) where T
 
-    if !(verbose in [0, 1, 2])
-        @warn("Incorrect verbose value input (should be 0, 1, or 2): switching to default value 1")
+    if !(verbose in [0, 1, 2, 3])
+        @warn("Incorrect verbose value input (should be 0, 1, 2 or 3): switching to default value 1")
         verbose = 1
     end
 

### 2019-10-24 src/random_seeds_lbfgs.jl

@@ -9,6 +9,7 @@ function estimate_statespace(model::StateSpaceModel{T}, filter_type::DataType,
     seeds         = Matrix{T}(undef, npsi, nseeds)
     loglikelihood = Vector{T}(undef, nseeds)
     psi           = Matrix{T}(undef, npsi, nseeds)
+    optseeds      = Vector{Optim.OptimizationResults}(undef, 0)
 
     # Initial conditions
     seedrange = collect(T, -1e3:0.1:1e3)
@@ -38,19 +39,23 @@ function estimate_statespace(model::StateSpaceModel{T}, filter_type::DataType,
                             LBFGS(), Optim.Options(f_tol = opt_method.f_tol, 
                                                    g_tol = opt_method.g_tol, 
                                                    iterations = opt_method.iterations,
-                                                   show_trace = (verbose == 2 ? true : false) ))
+                                                   show_trace = (verbose == 3 ? true : false) ))
         loglikelihood[iseed] = -optseed.minimum
         psi[:, iseed] = optseed.minimizer
+        push!(optseeds, optseed)
 
         print_loglikelihood(verbose, iseed, loglikelihood, t0)
     end
 
+    best_seed = argmax(loglikelihood)
+
     print_estimation_end(verbose, loglikelihood)
+    verbose >= 2 && println("Best seed optimization result: \n $(optseeds[best_seed])")
 
     # Put seeds in the model struct
     opt_method.seeds = seeds
 
-    bestpsi = psi[:, argmax(loglikelihood)]
+    bestpsi = psi[:, best_seed]
     H, Q    = statespace_covariance(bestpsi, model.dim.p, model.dim.r, filter_type)
 
     return StateSpaceCovariance(H, Q, filter_type)

### 2019-10-24 test/test_locallevel.jl

@@ -104,15 +104,15 @@ y = [0.0     # Deterministic local level generated time series
 
     @test isa(unimodel, StateSpaceModel)
     @test unimodel.mode == "time-invariant"
-    
-    ss1 = statespace(unimodel)
+
+    ss1 = statespace(unimodel; verbose = 2)
     @test ss1.filter_type <: KalmanFilter
     @test isa(ss1, StateSpace)
-    
+
     ss2 = statespace(unimodel; filter_type = SquareRootFilter{Float64})
     @test ss2.filter_type <: SquareRootFilter
     @test isa(ss2, StateSpace)
-    
+
     ss3 = statespace(unimodel; filter_type = UnivariateKalmanFilter{Float64})
     @test ss3.filter_type <: UnivariateKalmanFilter
     @test isa(ss3, StateSpace)

### 2019-10-24 benchmark/bench.jl
eliminate statespacecovariance

@@ -1,79 +1,39 @@
+push!(LOAD_PATH, "/home/guilhermebodin/Documents/Github/StateSpaceModels.jl/src")
 using StateSpaceModels, BenchmarkTools, CSV
 # Local level series
 locallevel_series = CSV.read("./benchmark/locallevel.csv"; header = false)
 verbose = 0
 
 locallevel = local_level(Matrix{Float64}(locallevel_series))
-bench_local_level = @benchmark statespace($locallevel, filter_type = $SquareRootFilter{Float64}, verbose = $verbose)
-# 9th August 2019
-# BenchmarkTools.Trial: 
-#   memory estimate:  135.70 MiB
-#   allocs estimate:  1613460
-#   --------------
-#   minimum time:     110.164 ms (11.13% GC)
-#   median time:      161.256 ms (11.65% GC)
-#   mean time:        161.873 ms (11.17% GC)
-#   maximum time:     206.068 ms (11.31% GC)
-#   --------------
-#   samples:          31
-#   evals/sample:     1
-
-bench_local_level = @benchmark statespace($locallevel, filter_type = $KalmanFilter{Float64}, verbose = $verbose)
-# 11th August 2019
-# BenchmarkTools.Trial: 
-#   memory estimate:  10.62 MiB
-#   allocs estimate:  92728
-#   --------------
-#   minimum time:     11.002 ms (0.00% GC)
-#   median time:      19.476 ms (7.28% GC)
-#   mean time:        20.071 ms (5.96% GC)
-#   maximum time:     45.660 ms (3.44% GC)
-#   --------------
-#   samples:          250
-#   evals/sample:     1
-
-bench_local_level = @benchmark statespace($locallevel, filter_type = $UnivariateKalmanFilter{Float64}, verbose = $verbose)
-# 22th August 2019
+locallevel.H[1] = 1
+locallevel.Q[1] = 1
+@benchmark StateSpaceModels.kalman_filter($locallevel)
+# 24th October 2019
 # BenchmarkTools.Trial:
-#   memory estimate:  7.09 MiB
-#   allocs estimate:  42550
-#   --------------
-#   minimum time:     5.756 ms (0.00% GC)
-#   median time:      10.094 ms (0.00% GC)
-#   mean time:        11.012 ms (12.64% GC)
-#   maximum time:     41.820 ms (24.80% GC)
-#   --------------
[…]

### 2019-10-24 src/StateSpaceModels.jl

@@ -9,6 +9,7 @@ export statespace, kfas, diagnostics
 include("prints.jl")
 include("structures.jl")
 include("utils.jl")
+include("model_unknowns.jl")
 include("models.jl")
 include("estimation.jl")
 include("random_seeds_lbfgs.jl")
@@ -35,24 +36,22 @@ function statespace(model::StateSpaceModel{T}; filter_type::DataType = KalmanFil
     print_header(verbose)
 
     # Maximum likelihood estimation
-    covariance = estimate_statespace(model, filter_type, optimization_method; verbose = verbose)
+    estimate_statespace!(model, filter_type, optimization_method; verbose = verbose)
 
     # Kalman filter and smoothing
-    filter_output, smoothed_state = kfas(model, covariance, filter_type)
+    filter_output, smoothed_state = kfas(model, filter_type)
 
     print_bottom(verbose)
 
-    return StateSpace{T}(model, filter_output, smoothed_state, covariance, filter_type, optimization_method)
+    return StateSpace{T}(model, filter_output, smoothed_state, filter_type, optimization_method)
 end
 
 """
-    kfas(model::StateSpaceModel, covariance::StateSpaceCovariance,
-         filter_type::DataType)
+    kfas(model::StateSpaceModel{T}, filter_type::DataType) where T
 
 Perform Kalman filter and smoother according to the chosen `filter_type`.
 """
-function kfas(model::StateSpaceModel{T}, covariance::StateSpaceCovariance{T},
-              filter_type::DataType) where T
+function kfas(model::StateSpaceModel{T}, filter_type::DataType) where T
     error(filter_type , " not implemented") # Returns an error if it cannot
                                             # find a specialized kfas
 end

### 2019-10-24 src/estimation.jl

@@ -30,16 +30,10 @@ function compute_log_likelihood_multivariate(n::Int, p::Int, v::Matrix{T}, F::Ar
     return log_likelihood
 end
 
-function get_log_likelihood_params(psitilde::Vector{T}, model::StateSpaceModel{T},
+function get_log_likelihood_params(model::StateSpaceModel{T},
                                    filter_type::DataType) where T
     error(filter_type , " not implemented") # Returns an error if it cannot 
                                             # find a specialized get_log_likelihood_params
-end
-
-function statespace_covariance(psi::Vector{T}, p::Int, r::Int,
-                               filter_type::DataType) where T
-    error(filter_type , " not implemented") # Returns an error if it cannot 
-                                            # find a specialized statespace_covariance
 end
 
 function valid_instants(model::StateSpaceModel{T}) where T
@@ -55,7 +49,7 @@ function valid_instants(model::StateSpaceModel{T}) where T
 end
 
 """
-    statespace_likelihood(psitilde::Vector{T}, model::StateSpaceModel{T}, 
+statespace_likelihood(psitilde::Vector{T}, model::StateSpaceModel, unknowns::Unknowns, 
                             valid_insts::Vector{Int}, filter_type::DataType) where T
 
 Compute log-likelihood concerning hyperparameter vector psitilde (``\\psi``)
@@ -63,10 +57,12 @@ Compute log-likelihood concerning hyperparameter vector psitilde (``\\psi``)
 Evaluate ``\\ell(\\psi;y_n)= -\\frac{np}{2}\\log2\\pi - \\frac{1}{2} \\sum_{t=1}^n \\log |F_t| - 
 \\frac{1}{2} \\sum_{t=1}^n v_t^{\\top} F_t^{-1} v_t ``
 """
-function statespace_likelihood(psitilde::Vector{T}, model::StateSpaceModel{T}, 
+function statespace_likelihood(psitilde::Vector{T}, model::StateSpaceModel, unknowns::Unknowns, 
                                valid_insts::Vector{Int}, filter_type::DataType) where T
+    # Fill all psitilde in the model
+    fill_model_with_psitilde!(model, psitilde, unknowns)
     # Calculate v and F 
-    v, F = get_log_likelihood_params(psitilde, model, filter_type)
+    v, F = get_log_likelihood_params(model, filter_type)
     # Compute log-likelihood based on v and F
     return compute_log_likelihood(model.dim.n, model.dim.p, v, F, valid_insts)
 end

### 2019-10-24 src/forecast.jl

@@ -9,8 +9,8 @@ at each time period.
 function forecast(ss::StateSpace{Typ}, N::Int) where Typ
 
     # Load estimated covariance matrices
-    H = ss.covariance.H
-    Q = ss.covariance.Q
+    H = ss.model.H
+    Q = ss.model.Q
 
     # Load system
     n, p, m, r = size(ss.model)
@@ -61,8 +61,8 @@ the number of series in the model, the number of steps ahead, and the number of 
 function simulate(ss::StateSpace{Typ}, N::Int, S::Int) where Typ
 
     # Load estimated covariance matrices
-    H = ss.covariance.H
-    Q = ss.covariance.Q
+    H = ss.model.H
+    Q = ss.model.Q
 
     # Load system
     n, p, m, r = size(ss.model)

### 2019-10-24 src/kalman.jl

@@ -1,9 +1,9 @@
 """
-    kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = 1e-5) where Typ <: Real
+    kalman_filter(model::StateSpaceModel; tol::Typ = 1e-5) where Typ
 
 Kalman filter with big Kappa initialization, i.e., initializing state variances as 1e6.
 """
-function kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ
+function kalman_filter(model::StateSpaceModel{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
     time_invariant = model.mode == "time-invariant"
 
@@ -33,7 +33,7 @@ function kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Ty
     fill_a1!(a)
     fill_P1!(P; bigkappa = Typ(1e6))
 
-    RQR = model.R * LinearAlgebra.BLAS.gemm('N', 'T', Typ(1.0), Q, model.R) # RQR = R Q R'
+    RQR = model.R * LinearAlgebra.BLAS.gemm('N', 'T', Typ(1.0), model.Q, model.R) # RQR = R Q R'
     # Kalman filter
     for t = 1:model.dim.n
         if t in model.missing_observations
@@ -55,7 +55,7 @@ function kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Ty
         else
             update_v!(v, model.y, model.Z, a, t) # v_t = y_t - Z_t * a_t
             update_ZP!(ZP, model.Z, P, t) # ZP = Z[:, :, t] * P[:, :, t]
-            update_F!(F, ZP, model.Z, H, t) # F_t = Z_t * P_t * Z_t + H
+            update_F!(F, ZP, model.Z, model.H, t) # F_t = Z_t * P_t * Z_t + H
             update_P_Ztransp_Finv!(P_Ztransp_invF, ZP, F, t) # P_Ztransp_invF   = ZP' * invF(F, t)
             update_K!(K, P_Ztransp_invF, model.T, t) # K_t = T * P_t * Z_t * F^-1_t
[…]

### 2019-10-24 src/model_unknowns.jl

@@ -1,0 +1,108 @@
+mutable struct Unknowns
+    n_unknowns::Int
+    unknown_indexes::Dict{String, Vector{CartesianIndex}}
+
+    function Unknowns(ssm::StateSpaceModel)
+        dict = Dict{String, Vector{CartesianIndex}}()
+        dict["Z"] = find_nans_Z(ssm.Z)
+        dict["T"] = find_nans(ssm.T)
+        dict["R"] = find_nans(ssm.R)
+        dict["sqrtH"] = find_nans_cov(ssm.H)
+        dict["sqrtQ"] = find_nans_cov(ssm.Q)
+
+        n_unknowns = num_unknowns(dict)
+
+        return new(n_unknowns, dict)
+    end
+end
+
+function find_nans(mat::Matrix{T}) where T
+    return findall(isnan, mat)
+end
+
+function find_nans_Z(Z::Array{T, 3}) where T
+    t_dim = size(Z, 3)
+    cart_idxs = Vector{Vector{CartesianIndex}}(undef, t_dim)
+    for i in 1:t_dim
+        cart_idxs[i] = findall(isnan, Z[:, :, i])
+    end
+
+    # Guarantee that the unkown is the same across all t
+    @assert all_indexes_are_equal(cart_idxs)
+
+    return cart_idxs[1]
+end
+
+function find_nans_cov(cov::Matrix{T}) where T
+    return findall(isnan, tril(cov))
+end
+
+function all_indexes_are_equal(cart_vec::Vector{Vector{CartesianIndex}})
+    for i in eachindex(cart_vec)
+        if cart_vec[i] != cart_vec[1]
+            return false
+        end
+    end
+    return true
+end
+
+function num_unknowns(dict::Dict{String, Vector{CartesianIndex}})
+    n = 0
+    for (k, v) in dict
+        n += length(v)
+    end
+    return n
+end
+
+function fill_model_with_psitilde!(model::StateSpaceModel{Typ}, psitilde::Vector{Typ}, unknowns::Unknowns) where Typ
+    # The order to fill the model is Z => T => R => sqrtH => sqrtQ
+    offset = 1
+
+    # Fill Z
+    for un_Z in unknowns.unknown_indexes["Z"]
+        for t in axes(model.Z, 3)
+            model.Z[un_Z, t] = psitilde[offset]
+        end
+        offset += 1
+    end
+
+    # Fill T
[…]

### 2019-10-24 src/random_seeds_lbfgs.jl

@@ -1,11 +1,18 @@
-function estimate_statespace(model::StateSpaceModel{T}, filter_type::DataType,
+function estimate_statespace!(model::StateSpaceModel{T}, filter_type::DataType,
                              opt_method::RandomSeedsLBFGS; verbose::Int = 1) where T
 
     nseeds = opt_method.nseeds                         
     nseeds += 1 # creating additional seed for degenerate cases
 
     # Initialization
-    npsi          = Int((1 + model.dim.r/model.dim.p)*(model.dim.p*(model.dim.p + 1)/2))
+    unknowns = Unknowns(model)
+
+    if unknowns.n_unknowns == 0
+        @warn("StateSpaceModel has no parameters to be estimated.")
+        return model
+    end
+ 
+    npsi          = unknowns.n_unknowns
     seeds         = Matrix{T}(undef, npsi, nseeds)
     loglikelihood = Vector{T}(undef, nseeds)
     psi           = Matrix{T}(undef, npsi, nseeds)
@@ -35,7 +42,7 @@ function estimate_statespace(model::StateSpaceModel{T}, filter_type::DataType,
 
     # Optimization
     for iseed = 1:nseeds
-        optseed = optimize(psitilde -> statespace_likelihood(psitilde, model, valid_insts, filter_type), seeds[:, iseed],
+        optseed = optimize(psitilde -> statespace_likelihood(psitilde, model, unknowns, valid_insts, filter_type), seeds[:, iseed],
                             LBFGS(), Optim.Options(f_tol = opt_method.f_tol, 
                                                    g_tol = opt_method.g_tol, 
                                                    iterations = opt_method.iterations,
@@ -54,9 +61,9 @@ function estimate_statespace(model::StateSpaceModel{T}, filter_type::DataType,
 
     # Put seeds in the model struct
     opt_method.seeds = seeds
-
+    # Take the best seed and fill the model with it
     bestpsi = psi[:, best_seed]
-    H, Q    = statespace_covariance(bestpsi, model.dim.p, model.dim.r, filter_type)
+    fill_model_with_psitilde!(model, bestpsi, unknowns)
 
-    return StateSpaceCovariance(H, Q, filter_type)
+    return 
 end

### 2019-10-24 src/sqrt_kalman.jl

@@ -1,9 +1,9 @@
 """
-    sqrt_kalman_filter(model::StateSpaceModel{Typ}, sqrtH::Matrix{Typ}, sqrtQ::Matrix{Typ}; tol::Typ = 1e-5) where Typ
+    sqrt_kalman_filter(model::StateSpaceModel{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
 Square-root Kalman filter with big Kappa initialization.
 """
-function sqrt_kalman_filter(model::StateSpaceModel{Typ}, sqrtH::Matrix{Typ}, sqrtQ::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ
+function sqrt_kalman_filter(model::StateSpaceModel{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
     time_invariant = model.mode == "time-invariant"
 
@@ -13,6 +13,12 @@ function sqrt_kalman_filter(model::StateSpaceModel{Typ}, sqrtH::Matrix{Typ}, sqr
     # Load system
     y = model.y
     Z, T, R = ztr(model)
+
+    ensure_pos_sym!(model.H)
+    ensure_pos_sym!(model.Q)
+
+    sqrtH = cholesky(model.H).L
+    sqrtQ = cholesky(model.Q).L
 
     # Predictive state and its sqrt-covariance
     a     = Matrix{Typ}(undef, n+1, m)
@@ -115,7 +121,7 @@ function filtered_state(model::StateSpaceModel{Typ}, sqrt_filter::SquareRootFilt
 end
 
 """
-    sqrt_smoother(model::StateSpaceModel{Typ}, sqrt_filter::SquareRootFilter{Typ}) where Typ
+    sqrt_smoother(model::StateSpaceModel, sqrt_filter::SquareRootFilter) where Typ
 
 Square-root smoother for state space model.
 """
@@ -198,49 +204,17 @@ function sqrt_smoother(model::StateSpaceModel{Typ}, sqrt_filter::SquareRootFilte
     return Smoother(alpha, V)
 end
 
+function get_log_likelihood_params(model::StateSpaceModel, filter_type::Type{SquareRootFilter{T}}) where T
 
-# All filters have to have implemented the following functions
-# *
-# *
-# *
-
-function statespace_covariance(psi::Vector{T}, p::Int, r::Int,
-                               filter_type::Type{SquareRootFilter{T}}) where T
-    # Observation sqrt-covariance matrix
[…]

### 2019-10-24 src/structures.jl

@@ -1,4 +1,4 @@
-export StateSpaceDimensions, StateSpaceModel, StateSpaceCovariance, SmoothedState,
+export StateSpaceDimensions, StateSpaceModel, SmoothedState,
         FilterOutput, StateSpace
 
 export KalmanFilter, SquareRootFilter, UnivariateKalmanFilter
@@ -112,16 +112,19 @@ Following the notation of on the book \"Time Series Analysis by State Space Meth
 * `Z` A ``p \\times m \\times n`` matrix
 * `T` A ``m \\times m`` matrix
 * `R` A ``m \\times r`` matrix
+* `H` A ``p \\times p`` matrix
+* `Q` A ``r \\times r`` matrix
 
 A `StateSpaceModel` object can be defined using `StateSpaceModel(y::Matrix{Typ}, Z::Array{Typ, 3}, T::Matrix{Typ}, R::Matrix{Typ})`.
-
 Alternatively, if `Z` is time-invariant, it can be input as a single ``p \\times m`` matrix.
 """
 struct StateSpaceModel{Typ <: Real}
-    y::Matrix{Typ} # observations
-    Z::Array{Typ, 3} # observation matrix
-    T::Matrix{Typ} # state matrix
-    R::Matrix{Typ} # state error matrix
+    y::Matrix{Typ} # Observations
+    Z::Array{Typ, 3} # Observation matrix
+    T::Matrix{Typ} # State matrix
+    R::Matrix{Typ} # State error matrix
+    H::Matrix{Typ} # Covariance matrix of the observation vector
+    Q::Matrix{Typ} # Covariance matrix of the state vector
     dim::StateSpaceDimensions
     missing_observations::Vector{Int}
     mode::String
@@ -137,7 +140,11 @@ struct StateSpaceModel{Typ <: Real}
             error("StateSpaceModel dimension mismatch")
         end
         dim = StateSpaceDimensions(ny, py, mr, rr)
-        new{Typ}(y, Z, T, R, dim, find_missing_observations(y), "time-variant")
+
+        # Build H and Q matrices with NaNs
+        H = build_H(dim.p, Typ)
+        Q = build_Q(dim.r, dim.p, Typ)
+        return new{Typ}(y, Z, T, R, H, Q, dim, find_missing_observations(y), "time-variant")
     end
 
     function StateSpaceModel(y::Matrix{Typ}, Z::Matrix{Typ}, T::Matrix{Typ}, R::Matrix{Typ}) where Typ <: Real
[…]

### 2019-10-24 src/univariate_kalman.jl

@@ -1,9 +1,9 @@
 """
-    univariate_kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ
+    kalman_filter(model::StateSpaceModel; tol::Typ = 1e-5) where Typ
 
 Kalman filter with big Kappa initialization, i.e., initializing state variances as 1e6.
 """
-function univariate_kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q::Matrix{Typ}; tol::Typ = Typ(1e-5)) where Typ
+function univariate_kalman_filter(model::StateSpaceModel{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
     time_invariant = model.mode == "time-invariant"
 
@@ -33,7 +33,7 @@ function univariate_kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q
     fill_a1!(a)
     fill_P1!(P; bigkappa = Typ(1e6))
 
-    mul!(RQR, model.R, LinearAlgebra.BLAS.gemm('N', 'T', Typ(1.0), Q, model.R)) # RQR = R Q R'
+    mul!(RQR, model.R, LinearAlgebra.BLAS.gemm('N', 'T', Typ(1.0), model.Q, model.R)) # RQR = R Q R'
     # Kalman filter
     for t = 1:model.dim.n
         if t in model.missing_observations
@@ -55,7 +55,7 @@ function univariate_kalman_filter(model::StateSpaceModel{Typ}, H::Matrix{Typ}, Q
         else
             update_v!(v, model.y, model.Z, a, t) # v_t = y_t - Z_t * a_t
             update_ZP!(ZP, model.Z, P, t) # ZP = Z[:, :, t] * P[:, :, t]
-            update_F!(F, ZP, model.Z, H, t) # F_t = Z_t * P_t * Z_t' + H
+            update_F!(F, ZP, model.Z, model.H, t) # F_t = Z_t * P_t * Z_t' + H
             update_P_Ztransp_Finv!(P_Ztransp_invF, ZP, F, t) # P_Ztransp_invF   = ZP' * invF(F, t)
             update_K!(K, P_Ztransp_invF, model.T, t) # K_t = T * P_t * Z_t * F^-1_t
[…]

### 2019-10-24 src/utils.jl

@@ -138,3 +138,14 @@ function Base.show(io::IO, model::StateSpaceModel)
     println("r = $(model.dim.r).")
     return nothing
 end
+
+function build_H(p::Int, T)
+    H = fill(NaN, p, p)
+    return T.(H)
+end
+
+function build_Q(r::Int, p::Int, T)
+    Q = kron(ones(p, p), Matrix{T}(I, Int(r/p), Int(r/p)))
+    Q[findall(isequal(1), Q)] .= NaN
+    return Q
+end

### 2019-10-24 test/test_lineartrend.jl

@@ -1,26 +1,29 @@
 @testset "Linear trend" begin
     @testset "Linear trend model" begin
         y = collect(1:15.0)
-        unimodel = linear_trend(y)
 
-        @test isa(unimodel, StateSpaceModel)
-        @test unimodel.mode == "time-invariant"
-                
-        ss1 = statespace(unimodel)
+        unimodel1 = linear_trend(y)
+        @test isa(unimodel1, StateSpaceModel)
+        @test unimodel1.mode == "time-invariant"
+        
+        ss1 = statespace(unimodel1)
 
+        unimodel2 = linear_trend(y)
         @test ss1.filter_type <: KalmanFilter
         @test isa(ss1, StateSpace)
         @test ss1.smoother.alpha[:, 2] ≈ ones(15) rtol = 1e-4
         compare_forecast_simulation(ss1, 20, 1000, 1e-3)
 
-        ss2 = statespace(unimodel; filter_type = SquareRootFilter{Float64})
+        ss2 = statespace(unimodel2; filter_type = SquareRootFilter{Float64})
 
         @test ss2.filter_type <: SquareRootFilter
         @test isa(ss2, StateSpace)
         @test ss2.smoother.alpha[:, 2] ≈ ones(15) rtol = 1e-4
         compare_forecast_simulation(ss2, 20, 1000, 1e-3)
 
-        ss3 = statespace(unimodel; filter_type = UnivariateKalmanFilter{Float64})
+        unimodel3 = linear_trend(y)
+
+        ss3 = statespace(unimodel3; filter_type = UnivariateKalmanFilter{Float64})
 
         @test ss3.filter_type <: UnivariateKalmanFilter
         @test isa(ss3, StateSpace)

### 2019-10-24 test/test_locallevel.jl

@@ -100,20 +100,24 @@ y = [0.0     # Deterministic local level generated time series
 8.366964896750876]
 
 @testset "Local level model with Kalman filter" begin
-    unimodel = local_level(y)
+    unimodel1 = local_level(y)
 
-    @test isa(unimodel, StateSpaceModel)
-    @test unimodel.mode == "time-invariant"
-
-    ss1 = statespace(unimodel; verbose = 2)
+    @test isa(unimodel1, StateSpaceModel)
+    @test unimodel1.mode == "time-invariant"
+    
+    ss1 = statespace(unimodel1)
     @test ss1.filter_type <: KalmanFilter
     @test isa(ss1, StateSpace)
 
-    ss2 = statespace(unimodel; filter_type = SquareRootFilter{Float64})
+    unimodel2 = local_level(y)
+    
+    ss2 = statespace(unimodel2; filter_type = SquareRootFilter{Float64})
     @test ss2.filter_type <: SquareRootFilter
     @test isa(ss2, StateSpace)
 
-    ss3 = statespace(unimodel; filter_type = UnivariateKalmanFilter{Float64})
+    unimodel3 = local_level(y)
+    
+    ss3 = statespace(unimodel3; filter_type = UnivariateKalmanFilter{Float64})
     @test ss3.filter_type <: UnivariateKalmanFilter
     @test isa(ss3, StateSpace)
 
@@ -122,16 +126,18 @@ y = [0.0     # Deterministic local level generated time series
     @test ss2.filter.a ≈ ss1.filter.a rtol = 1e-3
     @test ss3.filter.a ≈ ss1.filter.a rtol = 1e-3
 
-    unimodel32 = local_level(Float32.(y))
-    ss4 = statespace(unimodel32; filter_type = KalmanFilter{Float32})
+    unimodel32_1 = local_level(Float32.(y))
+    ss4 = statespace(unimodel32_1; filter_type = KalmanFilter{Float32})
     @test ss4.filter_type <: KalmanFilter
     @test isa(ss4, StateSpace)
-    
-    ss5 = statespace(unimodel32; filter_type = SquareRootFilter{Float32})
+
+    unimodel32_2 = local_level(Float32.(y))
+    ss5 = statespace(unimodel32_2; filter_type = SquareRootFilter{Float32})
[…]

### 2019-10-25 benchmark/bench.jl
fixed bug of H and Q exploding

@@ -1,4 +1,3 @@
-push!(LOAD_PATH, "/home/guilhermebodin/Documents/Github/StateSpaceModels.jl/src")
 using StateSpaceModels, BenchmarkTools, CSV
 # Local level series
 locallevel_series = CSV.read("./benchmark/locallevel.csv"; header = false)

### 2019-10-25 src/StateSpaceModels.jl

@@ -2,7 +2,7 @@ module StateSpaceModels
 
 using Optim, Distributions, LinearAlgebra, StaticArrays, Dates, Printf, StatsBase
 
-import Base: size, show
+import Base: size, show, copy
 
 export statespace, kfas, diagnostics
 

### 2019-10-25 src/model_unknowns.jl

@@ -56,41 +56,41 @@ end
 
 function fill_model_with_psitilde!(model::StateSpaceModel{Typ}, psitilde::Vector{Typ}, unknowns::Unknowns) where Typ
     # The order to fill the model is Z => T => R => sqrtH => sqrtQ
-    offset = 1
+    offset = 0
 
     # Fill Z
     for un_Z in unknowns.unknown_indexes["Z"]
+        offset += 1
         for t in axes(model.Z, 3)
             model.Z[un_Z, t] = psitilde[offset]
         end
-        offset += 1
     end
 
     # Fill T
     for un_T in unknowns.unknown_indexes["T"]
-        model.T[un_T] = psitilde[offset]
         offset += 1
+        model.T[un_T] = psitilde[offset]
     end
 
     # Fill R
     for un_R in unknowns.unknown_indexes["R"]
-        model.R[un_R] = psitilde[offset]
         offset += 1
+        model.R[un_R] = psitilde[offset]
     end
 
     # Fill sqrtH in the H matrix and do a gram (XX') after
     for un_H in unknowns.unknown_indexes["sqrtH"]
-        model.H[un_H] = psitilde[offset]
         offset += 1
+        model.H[un_H] = psitilde[offset]
     end
-    fill_H!(model, gram(tril(model.H)))
+    !isempty(unknowns.unknown_indexes["sqrtH"]) && fill_H!(model, gram(tril(model.H)))
 
     # Fill sqrtQ in the Q matrix and do a gram (XX') after
     for un_Q in unknowns.unknown_indexes["sqrtQ"]
-        model.Q[un_Q] = psitilde[offset]
         offset += 1
+        model.Q[un_Q] = psitilde[offset]
     end
-    fill_Q!(model, gram(tril(model.Q)))
+    !isempty(unknowns.unknown_indexes["sqrtQ"]) && fill_Q!(model, gram(tril(model.Q)))
 
     return 
 end

### 2019-10-25 src/random_seeds_lbfgs.jl

@@ -1,3 +1,18 @@
+"""
+TODO
+"""
+mutable struct RandomSeedsLBFGS{T <: Real} <: AbstractOptimizationMethod
+    f_tol::T
+    g_tol::T
+    iterations::Int
+    nseeds::Int
+    seeds::Array{T}
+
+    function RandomSeedsLBFGS(; f_tol::T = 1e-6, g_tol::T = 1e-6, iterations::Int = 10^5, nseeds::Int = 3) where T <: Real
+        return new{T}(f_tol, g_tol, 1e5, nseeds)
+    end
+end
+
 function estimate_statespace!(model::StateSpaceModel{T}, filter_type::DataType,
                              opt_method::RandomSeedsLBFGS; verbose::Int = 1) where T
 
@@ -42,16 +57,21 @@ function estimate_statespace!(model::StateSpaceModel{T}, filter_type::DataType,
 
     # Optimization
     for iseed = 1:nseeds
-        optseed = optimize(psitilde -> statespace_likelihood(psitilde, model, unknowns, valid_insts, filter_type), seeds[:, iseed],
-                            LBFGS(), Optim.Options(f_tol = opt_method.f_tol, 
-                                                   g_tol = opt_method.g_tol, 
-                                                   iterations = opt_method.iterations,
-                                                   show_trace = (verbose == 3 ? true : false) ))
-        loglikelihood[iseed] = -optseed.minimum
-        psi[:, iseed] = optseed.minimizer
-        push!(optseeds, optseed)
+        model_opt = copy(model)
+        try
+            optseed = optimize(psitilde -> statespace_likelihood(psitilde, model_opt, unknowns, valid_insts, filter_type), seeds[:, iseed],
+                                LBFGS(), Optim.Options(f_tol = opt_method.f_tol, 
+                                                       g_tol = opt_method.g_tol, 
+                                                       iterations = opt_method.iterations,
+                                                       show_trace = (verbose == 3 ? true : false) ))
+            loglikelihood[iseed] = -optseed.minimum
+            psi[:, iseed] = optseed.minimizer
+            push!(optseeds, optseed)
 
-        print_loglikelihood(verbose, iseed, loglikelihood, t0)
+            print_loglikelihood(verbose, iseed, loglikelihood, t0)
+        catch
+            println("seed diverged")
+        end
     end
 
     best_seed = argmax(loglikelihood)

### 2019-10-25 src/structures.jl

@@ -68,23 +68,6 @@ mutable struct Smoother{T <: Real} <: AbstractSmoother
     V::Array{T, 3} # variance of smoothed state
 end
 
-# Concrete structs for optimization methods
-"""
-TODO
-"""
-mutable struct RandomSeedsLBFGS{T <: Real} <: AbstractOptimizationMethod
-    f_tol::T
-    g_tol::T
-    iterations::Int
-    nseeds::Int
-    seeds::Array{T}
-
-    function RandomSeedsLBFGS(; f_tol::T = 1e-6, g_tol::T = 1e-6, iterations::Int = 10^5, nseeds::Int = 3) where T <: Real
-        return new{T}(f_tol, g_tol, 1e5, nseeds)
-    end
-end
-
-
 """
     StateSpaceDimensions
 
@@ -128,6 +111,20 @@ struct StateSpaceModel{Typ <: Real}
     dim::StateSpaceDimensions
     missing_observations::Vector{Int}
     mode::String
+
+    function StateSpaceModel(y::Matrix{Typ}, Z::Array{Typ, 3}, T::Matrix{Typ}, R::Matrix{Typ}, 
+                             H::Matrix{Typ}, Q::Matrix{Typ}, mode::String) where Typ <: Real
+        ny, py = size(y)
+        pz, mz, nz = size(Z)
+        mt1, mt2 = size(T)
+        mr, rr = size(R)
+        if !((mz == mt1 == mt2 == mr) && (pz == py))
+            error("StateSpaceModel dimension mismatch")
+        end
+        dim = StateSpaceDimensions(ny, py, mr, rr)
+
+        return new{Typ}(y, Z, T, R, H, Q, dim, find_missing_observations(y), mode)
+    end
 
     function StateSpaceModel(y::Matrix{Typ}, Z::Array{Typ, 3}, T::Matrix{Typ}, R::Matrix{Typ}) where Typ <: Real
 

### 2019-10-25 src/utils.jl

@@ -1,4 +1,4 @@
-export size, ztr
+export size, ztr, copy
 
 """
     size(model::StateSpaceModel)
@@ -137,6 +137,10 @@ function Base.show(io::IO, model::StateSpaceModel)
     println("m = $(model.dim.m),")
     println("r = $(model.dim.r).")
     return nothing
+end
+
+function copy(ssm::StateSpaceModel)
+    return StateSpaceModel(copy(ssm.y), copy(ssm.Z), copy(ssm.T), copy(ssm.R), copy(ssm.H), copy(ssm.Q), ssm.mode)
 end
 
 function build_H(p::Int, T)

### 2019-10-25 test/test_locallevel.jl

@@ -180,3 +180,36 @@ end
     @test ss.model.H ≈ [1.0 0.5; 0.5 1.0] rtol = 1e-1
     @test ss.model.Q ≈ [1.0 0.5; 0.5 1.0] rtol = 1e-1
 end
+
+# push!(LOAD_PATH, "/home/guilhermebodin/Documents/Github/StateSpaceModels.jl/src")
+# using StateSpaceModels
+
+# Z = [1.0][:, :]
+# T = [1.0][:, :]
+# R = [1.0][:, :]
+# ssm = StateSpaceModel(y[:, :], Z, T, R)
+# ssm.H[1] = 1.42
+# ssm.Q[1] = NaN
+
+# opt_method = RandomSeedsLBFGS()
+# ss = statespace(ssm; optimization_method = opt_method)
+# ss.model.H
+# ss.model.Q
+
+# unimodel1 = local_level(y)
+# ss1 = statespace(unimodel1)
+
+# Z = [NaN][:, :]
+# T = [1.0][:, :]
+# R = [1.0][:, :]
+# ssm = StateSpaceModel(y[:, :], Z, T, R)
+# ssm.H[1] = NaN
+# ssm.Q[1] = 1.24
+
+# opt_method = RandomSeedsLBFGS(nseeds = 10)
+# ss = statespace(ssm; optimization_method = opt_method)
+# ss.model.Z
+# ss.model.T
+# ss.model.R
+# ss.model.H
+# ss.model.Q

### 2019-10-25 src/estimation.jl
added statespace_recursion function

@@ -60,7 +60,7 @@ Evaluate ``\\ell(\\psi;y_n)= -\\frac{np}{2}\\log2\\pi - \\frac{1}{2} \\sum_{t=1}
 function statespace_likelihood(psitilde::Vector{T}, model::StateSpaceModel, unknowns::Unknowns, 
                                valid_insts::Vector{Int}, filter_type::DataType) where T
     # Fill all psitilde in the model
-    fill_model_with_psitilde!(model, psitilde, unknowns)
+    fill_model_with_parameters!(model, psitilde, unknowns)
     # Calculate v and F 
     v, F = get_log_likelihood_params(model, filter_type)
     # Compute log-likelihood based on v and F

### 2019-10-25 src/kalman.jl

@@ -1,5 +1,5 @@
 """
-    kalman_filter(model::StateSpaceModel; tol::Typ = 1e-5) where Typ
+    kalman_filter(model::StateSpaceModel{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
 Kalman filter with big Kappa initialization, i.e., initializing state variances as 1e6.
 """
@@ -73,7 +73,7 @@ function kalman_filter(model::StateSpaceModel{Typ}; tol::Typ = Typ(1e-5)) where 
 end
 
 """
-    smoother(model::StateSpaceModel, kfilter::KalmanFilter) where Typ
+    smoother(model::StateSpaceModel{Typ}, kfilter::KalmanFilter{Typ}) where Typ
 
 Smoother for state-space model.
 """

### 2019-10-25 src/model_unknowns.jl

@@ -27,7 +27,7 @@ function find_nans_Z(Z::Array{T, 3}) where T
         cart_idxs[i] = findall(isnan, Z[:, :, i])
     end
 
-    # Guarantee that the unkown is the same across all t
+    # Guarantee that the unknown is the same across all t
     @assert all_indexes_are_equal(cart_idxs)
 
     return cart_idxs[1]
@@ -54,7 +54,7 @@ function num_unknowns(dict::Dict{String, Vector{CartesianIndex}})
     return n
 end
 
-function fill_model_with_psitilde!(model::StateSpaceModel{Typ}, psitilde::Vector{Typ}, unknowns::Unknowns) where Typ
+function fill_model_with_parameters!(model::StateSpaceModel{Typ}, psitilde::Vector{Typ}, unknowns::Unknowns) where Typ
     # The order to fill the model is Z => T => R => sqrtH => sqrtQ
     offset = 0
 

### 2019-10-25 src/random_seeds_lbfgs.jl

@@ -83,7 +83,7 @@ function estimate_statespace!(model::StateSpaceModel{T}, filter_type::DataType,
     opt_method.seeds = seeds
     # Take the best seed and fill the model with it
     bestpsi = psi[:, best_seed]
-    fill_model_with_psitilde!(model, bestpsi, unknowns)
+    fill_model_with_parameters!(model, bestpsi, unknowns)
 
     return 
 end

### 2019-10-25 src/univariate_kalman.jl

@@ -1,5 +1,5 @@
 """
-    kalman_filter(model::StateSpaceModel; tol::Typ = 1e-5) where Typ
+    univariate_kalman_filter(model::StateSpaceModel{Typ}; tol::Typ = Typ(1e-5)) where Typ
 
 Kalman filter with big Kappa initialization, i.e., initializing state variances as 1e6.
 """
@@ -73,7 +73,7 @@ function univariate_kalman_filter(model::StateSpaceModel{Typ}; tol::Typ = Typ(1e
 end
 
 """
-    univariate_smoother(model::StateSpaceModel, kfilter::UnivariateKalmanFilter) where Typ
+    univariate_smoother(model::StateSpaceModel{Typ}, kfilter::UnivariateKalmanFilter{Typ}) where Typ
 
 Smoother for state-space model.
 """

### 2019-10-25 src/utils.jl

@@ -1,4 +1,4 @@
-export size, ztr, copy
+export size, ztr, copy, statespace_recursion
 
 """
     size(model::StateSpaceModel)
@@ -16,6 +16,42 @@ Return the state space model arrays `Z`, `T` and `R` of the `StateSpaceModel`
 """
 function ztr(model::StateSpaceModel)
     return model.Z, model.T, model.R
+end
+
+"""
+    statespace_recursion(model::StateSpaceModel{Typ}, N::Int, initial_a::Matrix{Typ}) where Typ
+
+Runs the recursion of a state space model ignoring the y.
+"""
+function statespace_recursion(model::StateSpaceModel{Typ}, initial_a::Matrix{Typ}) where Typ
+
+    # If model has any unknows is not possible to perform the recursion
+    unknowns = Unknowns(model)
+    if unknowns.n_unknowns != 0
+        error("StateSpaceModel has unkown parameters.")
+    end
+
+    # Save the distributions H and Q
+    dist_H = MvNormal(model.H)
+    dist_Q = MvNormal(model.Q)
+
+    n, p = size(model.y)
+
+    if size(initial_a, 2) != p
+        error("intial_a must be a 1 by $p matrix.")
+    end
+
+    y = Matrix{Float64}(undef, n, p)
+    α = Matrix{Float64}(undef, n, p)
+    α[1, :] = initial_a
+
+    for t = 1:n-1
+        y[t, :]   = model.Z[:, :, t]*α[t, :] + rand(dist_H)
+        α[t+1, :] = model.T*α[t, :] + model.R*rand(dist_Q)
+    end
+    y[n, :] = model.Z[:, :, n]*α[n, :] + rand(dist_H)
+
+    return y, α
 end
 
 

### 2019-10-25 test/test_locallevel.jl

@@ -151,28 +151,19 @@ y = [0.0     # Deterministic local level generated time series
 end
 
 @testset "Correlated multivariate test" begin
-    # Define error covariances
-    dist_H = MvNormal([1.0 0.5;
-                        0.5 1.0])
-
-    dist_Q = MvNormal([1.0 0.5;
-                        0.5 1.0])
-
     # Sample two observation series from defined errors
     Random.seed!(1)
     n = 1000
     p = 2
-    y = Matrix{Float64}(undef, n, p)
-    α = Matrix{Float64}(undef, n, p)
+    α = Matrix{Float64}(undef, 1, p)
+    y = ones(n, p)
 
     model = local_level(y)
+    model.Q .= [1.0 0.5; 0.5 1.0]
+    model.H .= [1.0 0.5; 0.5 1.0]
     α[1, :] = [1.0, 1.0]
 
-    for t = 1:n-1
-        y[t, :]   = model.Z[:, :, t]*α[t, :] + rand(dist_H)
-        α[t+1, :] = model.T*α[t, :] + model.R*rand(dist_Q)
-    end
-    y[n, :] = model.Z[:, :, n]*α[n, :] + rand(dist_H)
+    y, α_n = statespace_recursion(model, α)
 
     model = local_level(y)
     ss = statespace(model)
@@ -181,8 +172,9 @@ end
     @test ss.model.Q ≈ [1.0 0.5; 0.5 1.0] rtol = 1e-1
 end
 
-# push!(LOAD_PATH, "/home/guilhermebodin/Documents/Github/StateSpaceModels.jl/src")
-# using StateSpaceModels
+@testset "Unknowns in Z, T and R" begin
+    #TODO
+end
 
 # Z = [1.0][:, :]
 # T = [1.0][:, :]

### 2019-10-25 test/test_utils.jl

@@ -61,6 +61,10 @@
         @test show(model) == nothing
         @test show(ss) == nothing
     end
+
+    @testset "statespace_recursion" begin
+        #TODO
+    end
 end
 
 function compare_forecast_simulation(ss::StateSpace, N::Int, S::Int, rtol::Float64)

### 2020-02-18 examples/vehicle_tracking.jl
update gif and add it to README

@@ -1,3 +1,4 @@
+push!(LOAD_PATH, "/home/guilhermebodin/Documents/Github/StateSpaceModels.jl/src")
 using StateSpaceModels, Distributions, LinearAlgebra, Plots, Random
 
 # Number of observations
@@ -44,6 +45,10 @@ model = StateSpaceModel(y, Z, T, R)
 # Estimate vehicle speed and position
 ss = statespace(model)
 
-plot(y[:, 1], y[:, 2], label="Measured position", line=:scatter, lw=2, markeralpha=0.2, color=:black, title="Vehicle tracking")
-plot!(α[:, 1], α[:, 3], label="True position", lw=3, color=:indianred)
-plot!(ss.smoother.alpha[:, 1], ss.smoother.alpha[:, 3], label="Estimated position", lw=2, color=:forestgreen)
+anim = @animate for i in 1:n
+    plot(y[1:i, 1], y[1:i, 2], label="Measured position", line=:scatter, lw=2, markeralpha=0.2, color=:black, title="Vehicle tracking")
+    plot!(α[1:i, 1], α[1:i, 3], label="True position", lw=3, color=:indianred)
+    plot!(ss.filter.a[2:i+1, 1], ss.filter.a[2:i+1, 3], label="Estimated position", lw=2, color=:forestgreen)
+end
+
+gif(anim, "vehicle_tracking.gif", fps = 15)

### 2020-02-18 examples/vehicle_tracking.jl
Update gif (#148)

@@ -52,7 +52,7 @@ ss.model.Q
 anim = @animate for i in 1:n
     plot(y[1:i, 1], y[1:i, 2], label="Measured position", line=:scatter, lw=2, markeralpha=0.2, color=:black, title="Vehicle tracking")
     plot!(α[1:i, 1], α[1:i, 3], label="True position", lw=3, color=:indianred)
-    plot!(ss.smoother.alpha[1:i, 1], ss.smoother.alpha[1:i, 3], label="Estimated position", lw=2, color=:forestgreen)
+    plot!(ss.filter.a[2:i+1, 1], ss.filter.a[2:i+1, 3], label="Estimated position", lw=2, color=:forestgreen)
 end
 
 gif(anim, "vehicle_tracking.gif", fps = 15)

### 2020-02-20 paper/paper.tex
Update paper (#151)

@@
 The contributions brought by these packages notwithstanding, there are currently no state-space model packages aimed at time-series analysis in Julia.
-Furthermore, some have not been updated for a long time and are not functional in Julia 1.0.
 Thus, the main objectives of \texttt{StateSpaceModels.jl} \cite{statespacemodels} are:
@@
 \noindent where \texttt{y} is an observation vector in the univariate case or a matrix in column-wise fashion in the multivariate case, i.e., each column representing a variable and each line representing a time period.
+
+While very simple, the local level model can be successfully applied to a wide array of problems wherein one wants to estimate some magnitude that varies over time.
+For instance, in \cite{koopman2000stamp}, an illustrative example utilizes water inflow data from the river Nile.
 %
@@
 \eqref{eq:lineartrend1}--\eqref{eq:lineartrend3}.
+Note that it differs from the local level model due to the inclusion of the slope $\nu_{t}$ in \eqref{eq:lineartrend2}.
 A linear trend \texttt{StateSpaceModel} can be created using:
@@
 %
-In this model, there is a stochastic trend that contains a slope component, similarly to the linear trend model.
-Furthermore, a stochastic seasonal component is modeled so that its sum over the seasonality period $s$ is equal to zero except for a noise factor.
+In this model, there is a stochastic trend that contains a slope component, exactly as the linear trend model.
+Furthermore, a stochastic seasonal component is modeled such that its sum over the seasonality period $s$ is equal to zero except for a noise factor.
 A basic structural \texttt{StateSpaceModel} can be created using:
@@
 \end{lstlisting}
+
+The idea of the seasonal component is to account for effects that have a periodic nature, such as monthly temperature over each year.
 
@@
 
-The filtering step derives the predictive and filtered states as well as their covariance matrices at every time period via the Kalman filter. The predictive state and its variance are given by
+The filtering step computes the predictive and filtered states as well as their covariance matrices at every time period via the Kalman filter. The predictive state and its variance are given by
 %
@@
 %
-which depends on the innovations and their covariance matrix at each time period. Therefore, at each optimization iteration, the Kalman filter needs to be executed.
+which depends on the innovations and their covariance matrix at each time period.
+Therefore, at each iteration of the optimization, the Kalman filter needs to be executed to compute the objective function.
 
@@
 
-As a way of evaluating the model specification and estimation, the package contains statistical diagnostics that can be conducted over the residuals.
+As a way of evaluating the model specification and estimation, the package contains basic statistical diagnostics that can be conducted over the residuals.
 The available diagnostics are the Jarque-Bera normality test \cite{jarque1980efficient}, the Ljung-Box independence test \cite{box1970distribution}, and an homoscedasticity test \cite{durbin2012time}.
@@
 As a first example, let us use the classical monthly airline passengers time series.
-In order to avoid multiplicative effects, we use the well-known approach of taking the log of the series.
+In order to avoid multiplicative effects, we use the well-known approach of taking the log of the series \cite{lutkepohl2012role}.
 Fig.
@@
 
-By estimating a structural model, we can analyzed the individual components of the series, such as trend and seasonality.
+By estimating a structural model, we can analyze the individual components of the series, such as trend and seasonality.
 These components are presented in Fig.

### 2020-02-25 paper/paper.tex
Add reference to repo

@@
 We utilize \texttt{StateSpaceModels.jl} to tackle the presented problems and provide the results.
-Furthermore, for the sake of reproducibility, the code of some of the examples is in the folder \texttt{examples} in the \texttt{StateSpaceModels.jl} repository.
+Furthermore, for the sake of reproducibility, the code of some of the examples is in the folder \texttt{examples} in the \texttt{StateSpaceModels.jl} repository \cite{statespacemodels}.
 
@@
 	\includegraphics[width=\columnwidth]{Figures/log_airline_forecast.pdf}
-	\caption{Log-airline passengers historical data and forecasting.}
+	\caption{Log-airline passengers historical data and forecast for the following two years.}
 	\label{fig:log_airline_forecast}
